---
title: "Compile Decoy Intron Data (v3 — overlap first, then bash features)"
output: html_document
---

<!--
  v3 workflow (two parts):
  Part 1: Load original slop24 BED, overlap with PSI introns, strand-aware distance, write table + BED stubs for bash.
  Between: Run phastCons (intron segments) and MaxEntScan (decoy + canonical) in bash; save TSV/BED outputs.
  Part 2: Reload Part-1 table, merge scores, plot, write final TSV.

  Input slop24: original 7-column BED (column 7 dropped on load).
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-libraries}
library(data.table)
library(GenomicRanges)
library(ggplot2)
```

# Part 1 — Load slop24, overlap introns, write intermediate table

```{r paths-v3}
slop24_bed <- "../data/Decoys/Decoys_proteincoding_slop24.bed"
psi_table <- "../reference/PSI_TABLE-hg38.tab.gz"
out_dir <- "../results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

step1_tsv <- file.path(out_dir, "decoy_intron_overlap_step1.tsv")
```

## Load original slop24 BED

```{r load-slop24-v3}
bed <- fread(
  slop24_bed,
  sep = "\t",
  header = FALSE,
  col.names = c("chr", "start", "end", "decoyID", "SpliceAI_score", "strand", "counts")
)
bed[, counts := NULL]
bed[, decoyID := paste0(decoyID, "_", start)]

cat("Rows:", nrow(bed), " Unique decoyID:", uniqueN(bed$decoyID),
    " Duplicates:", sum(duplicated(bed$decoyID)), "\n")
```

## PSI table: HsaIN only, drop `-Q` columns, parse intron coordinates

```{r load-psi-v3}
psi <- fread(psi_table, sep = "\t", header = TRUE)
psi <- psi[grepl("^HsaIN", EVENT)]
q_cols <- grep("-Q$", names(psi), value = TRUE)
if (length(q_cols) > 0) psi[, (q_cols) := NULL]

coord_parts <- tstrsplit(psi$COORD, ":", fixed = TRUE)
range_parts <- tstrsplit(coord_parts[[2]], "-", fixed = TRUE)
psi[, chr := coord_parts[[1]]]
psi[, intron_start := as.integer(range_parts[[1]])]
psi[, intron_end := as.integer(range_parts[[2]])]
psi <- psi[!is.na(chr) & !is.na(intron_start) & !is.na(intron_end)]
cat("PSI rows after filters:", nrow(psi), "\n")
```

## Overlap decoys with introns (shortest intron per decoy)

```{r overlap-v3}
decoy_gr <- GRanges(
  seqnames = bed$chr,
  ranges = IRanges(start = bed$start + 1, end = bed$end),
  strand = bed$strand
)
psi_gr <- GRanges(
  seqnames = psi$chr,
  ranges = IRanges(start = psi$intron_start, end = psi$intron_end)
)
hits <- findOverlaps(decoy_gr, psi_gr, ignore.strand = TRUE)
cat("Overlap pairs:", length(hits), "\n")

bed_hit <- bed[queryHits(hits)]
psi_hit <- psi[subjectHits(hits)]
overlap_cols <- intersect(names(bed_hit), names(psi_hit))
if (length(overlap_cols) > 0) setnames(psi_hit, overlap_cols, paste0("psi_", overlap_cols))
merged <- cbind(bed_hit, psi_hit)

merged[, intron_length := intron_end - intron_start + 1L]
setorder(merged, decoyID, intron_length)
merged <- merged[!duplicated(decoyID)]
cat("Rows after shortest-intron selection:", nrow(merged), "\n")
```

## Strand-aware distance from canonical 5' splice site

```{r distance-v3}
merged[, decoy_coord_1based := start + 1L]
merged[, decoy_distance_from_canonical_5ss := ifelse(
  strand == "+",
  decoy_coord_1based - intron_start,
  intron_end - decoy_coord_1based
)]
```

## Write Part-1 output (main table for bash + Part 2)

```{r write-step1}
fwrite(merged, step1_tsv, sep = "\t", quote = FALSE)
cat("Wrote:", step1_tsv, " rows:", nrow(merged), " cols:", ncol(merged), "\n")
```

## Optional: BED files for bash (phastCons on intron; MaxEnt on decoy / canonical)

These use the **intron interval** for phastCons averaging (full segment), and decoy vs canonical 5'SS coordinates for MaxEnt (same logic as earlier pipeline).

```{r write-bash-input-beds}
intron_bed <- file.path(out_dir, "v3_intron_segments_for_phastcons.bed")
decoy_bed <- file.path(out_dir, "v3_decoy_coords_for_maxent.bed")
canonical_bed <- file.path(out_dir, "v3_canonical_coords_for_maxent.bed")

fwrite(
  merged[, .(chr, intron_start - 1L, intron_end, decoyID, SpliceAI_score, strand)],
  intron_bed,
  sep = "\t",
  col.names = FALSE
)
fwrite(
  merged[, .(chr, start, end, decoyID, SpliceAI_score, strand)],
  decoy_bed,
  sep = "\t",
  col.names = FALSE
)
canonical_dt <- merged[, .(
  chr = chr,
  start = ifelse(strand == "+", intron_start - 1L, intron_end - 1L),
  end = ifelse(strand == "+", intron_start, intron_end),
  decoyID = decoyID,
  SpliceAI_score = SpliceAI_score,
  strand = strand
)]
fwrite(canonical_dt, canonical_bed, sep = "\t", col.names = FALSE)

cat("Wrote:\n ", intron_bed, "\n ", decoy_bed, "\n ", canonical_bed, "\n")
```

---

# Between Part 1 and Part 2 — bash intermediate (phastCons + MaxEnt)

**Stop here after knitting Part 1** (or run chunks through `write-bash-input-beds`), then run the following in the shell from `hnRNPH1_IR_MAF` (adjust paths if needed).

### phastCons (mean over each **intron** segment)

Use `scripts/extract_phastcons_scores.sh` (or direct `bigWigAverageOverBed`) with:

- BED: `results/v3_intron_segments_for_phastcons.bed` (column 4 = `decoyID`, unique per row)
- BigWigs: `reference/hg38.phastCons100way.bw` and `reference/hg38.phastCons470way.bw`

The script prints `decoyID<TAB>mean` to stdout. Save:

- `results/v3_phastcons100_by_decoy.tsv` — two columns, no header: `decoyID`, `phastCons100_mean`
- `results/v3_phastcons470_by_decoy.tsv` — `decoyID`, `phastCons470_mean`

Example:

```bash
bash scripts/extract_phastcons_scores.sh reference/hg38.phastCons100way.bw results/v3_intron_segments_for_phastcons.bed > results/v3_phastcons100_by_decoy.tsv
bash scripts/extract_phastcons_scores.sh reference/hg38.phastCons470way.bw results/v3_intron_segments_for_phastcons.bed > results/v3_phastcons470_by_decoy.tsv
```

### MaxEntScan (decoy site vs canonical 5'SS)

Use `scripts/run_maxentscan_decoys.sh` and `scripts/run_maxentscan_canonical.sh` with genome FASTA as third argument, e.g.:

```bash
bash scripts/run_maxentscan_decoys.sh results/v3_decoy_coords_for_maxent.bed results/v3_maxent_decoy.bed ../../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa
bash scripts/run_maxentscan_canonical.sh results/v3_canonical_coords_for_maxent.bed results/v3_maxent_canonical.bed ../../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa
```

Scores are appended as the **last column** of each output BED (column 4 = `decoyID`).

---

# Part 2 — Reload step-1 table, merge scores, plot, final export

```{r paths-part2}
out_dir <- "../results"
step1_tsv <- file.path(out_dir, "decoy_intron_overlap_step1.tsv")
ph100_tsv <- file.path(out_dir, "v3_phastcons100_by_decoy.tsv")
ph470_tsv <- file.path(out_dir, "v3_phastcons470_by_decoy.tsv")
maxent_decoy_bed <- file.path(out_dir, "v3_maxent_decoy.bed")
maxent_canonical_bed <- file.path(out_dir, "v3_maxent_canonical.bed")
final_tsv <- file.path(out_dir, "decoy_intron_features_v3_final.tsv")
```

```{r reload-step1}
if (!file.exists(step1_tsv)) {
  stop("Missing Part-1 table: ", step1_tsv, " — knit Part 1 first.")
}
merged <- fread(step1_tsv, sep = "\t", header = TRUE)
cat("Reloaded rows:", nrow(merged), " cols:", ncol(merged), "\n")
```

```{r merge-phastcons}
if (file.exists(ph100_tsv)) {
  ph100 <- fread(ph100_tsv, sep = "\t", header = FALSE, col.names = c("decoyID", "phastCons100_mean"))
  merged <- merge(merged, ph100, by = "decoyID", all.x = TRUE)
} else {
  merged[, phastCons100_mean := NA_real_]
  cat("Note: missing ", ph100_tsv, " — phastCons100 not merged.\n")
}
if (file.exists(ph470_tsv)) {
  ph470 <- fread(ph470_tsv, sep = "\t", header = FALSE, col.names = c("decoyID", "phastCons470_mean"))
  merged <- merge(merged, ph470, by = "decoyID", all.x = TRUE)
} else {
  merged[, phastCons470_mean := NA_real_]
  cat("Note: missing ", ph470_tsv, " — phastCons470 not merged.\n")
}
```

```{r merge-maxent}
if (file.exists(maxent_decoy_bed)) {
  raw <- fread(maxent_decoy_bed, sep = "\t", header = FALSE)
  sc <- unique(
    data.table(decoyID = raw[[4]], MaxEnt_decoy = as.numeric(raw[[ncol(raw)]])),
    by = "decoyID"
  )
  merged <- merge(merged, sc, by = "decoyID", all.x = TRUE)
} else {
  merged[, MaxEnt_decoy := NA_real_]
  cat("Note: missing ", maxent_decoy_bed, "\n")
}
if (file.exists(maxent_canonical_bed)) {
  raw <- fread(maxent_canonical_bed, sep = "\t", header = FALSE)
  sc <- unique(
    data.table(decoyID = raw[[4]], MaxEnt_canonical = as.numeric(raw[[ncol(raw)]])),
    by = "decoyID"
  )
  merged <- merge(merged, sc, by = "decoyID", all.x = TRUE)
} else {
  merged[, MaxEnt_canonical := NA_real_]
  cat("Note: missing ", maxent_canonical_bed, "\n")
}
```

## PSI count (columns with value >= 10)

```{r psi-count-v3}
meta_cols <- c("GENE", "EVENT", "COORD", "LENGTH", "FullCO", "COMPLEX",
               "chr", "intron_start", "intron_end", "intron_length")
feat <- c(
  "decoy_coord_1based", "decoy_distance_from_canonical_5ss",
  "start", "end", "decoyID", "SpliceAI_score", "strand",
  "MaxEnt_decoy", "MaxEnt_canonical", "phastCons470_mean", "phastCons100_mean"
)
psi_cols <- setdiff(names(merged), c(meta_cols, feat, "chr.1"))
psi_num <- psi_cols[sapply(merged[, ..psi_cols], is.numeric)]
merged[, psi_count_ge10 := rowSums(.SD >= 10, na.rm = TRUE), .SDcols = psi_num]
```

```{r reorder-v3}
bed_cols <- c("chr", "start", "end", "decoyID", "SpliceAI_score", "strand")
pref <- c(
  "decoy_coord_1based", "decoy_distance_from_canonical_5ss",
  "MaxEnt_decoy", "MaxEnt_canonical",
  "phastCons470_mean", "phastCons100_mean", "psi_count_ge10"
)
pref <- pref[pref %in% names(merged)]
setcolorder(merged, c(bed_cols, pref, setdiff(names(merged), c(bed_cols, pref))))
```

## Plot: log distance vs PSI-count, size = SpliceAI, color = phastCons470

```{r plot-v3}
plot_df <- merged[!is.na(decoy_distance_from_canonical_5ss) & !is.na(psi_count_ge10)]
plot_df[, log_distance_from_5ss := log10(abs(decoy_distance_from_canonical_5ss) + 1)]
highlight_id <- "HNRNPH1_179620582"

p <- ggplot(
  plot_df,
  aes(
    x = log_distance_from_5ss,
    y = psi_count_ge10,
    size = SpliceAI_score,
    color = phastCons470_mean
  )
) +
  geom_point(alpha = 0.5) +
  geom_point(data = plot_df[decoyID == highlight_id], color = "red", alpha = 0.9, stroke = 1.2) +
  geom_text(
    data = plot_df[decoyID == highlight_id],
    aes(label = decoyID),
    color = "red",
    vjust = -0.8,
    size = 3,
    show.legend = FALSE
  ) +
  labs(
    title = "Log distance from canonical 5'SS vs PSI columns >= 10",
    x = "log10(|distance| + 1)",
    y = "Count of PSI columns >= 10",
    size = "SpliceAI score",
    color = "phastCons 470-way (intron mean)"
  ) +
  scale_color_viridis_c(option = "C", na.value = "grey80") +
  theme_bw()

print(p)
```

```{r write-final-v3}
fwrite(merged, final_tsv, sep = "\t", quote = FALSE)
cat("Final table:", final_tsv, "\n")
```
