---
title: "Compile Decoy Intron Data From Slop24 (v2 Inline)"
output: html_document
---

This workflow expects a precomputed slop24 BED with phastCons columns:
`../data/Decoys/Decoys_proteincoding_slop24_with_phastcons.bed`.

The phastCons values were computed outside this Rmd using
`scripts/extract_phastcons_scores.sh` on a temporary BED where column 4 was made unique
as `GENE_start` (required by `bigWigAverageOverBed`), then merged back onto the original slop24 BED:

- phastCons 470-way from `reference/hg38.phastCons470way.bw` -> appended as column 9
- phastCons 100-way from `reference/hg38.phastCons100way.bw` -> appended as column 8

Final BED column layout used here:
1=`chr`, 2=`start`, 3=`end`, 4=`gene`, 5=`SpliceAI_score`, 6=`strand`,
7=`counts` (dropped on load), 8=`phastCons100_mean`, 9=`phastCons470_mean`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-libraries}
library(data.table)
library(GenomicRanges)
library(ggplot2)
library(VarCon)
```

## Inputs and Paths

```{r paths}
slop24_bed <- "../data/Decoys/Decoys_proteincoding_slop24_with_phastcons.bed"
psi_table <- "../reference/PSI_TABLE-hg38.tab.gz"
genome_fa <- "../../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa"

out_dir <- "../results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

genome_fa <- normalizePath(genome_fa, mustWork = TRUE)
```

## Step 0: Load slop24 BED

```{r load-slop24}
bed <- fread(
  slop24_bed,
  sep = "\t",
  header = FALSE,
  col.names = c(
    "chr", "start", "end", "decoyID", "SpliceAI_score", "strand",
    "counts", "phastCons100_mean", "phastCons470_mean"
  )
)
bed[, counts := NULL]
bed[, decoyID := paste0(decoyID, "_", start)]

cat("Loaded slop24 BED rows:", nrow(bed), "\n")
cat("Unique decoy IDs:", uniqueN(bed$decoyID), "\n")
cat("Duplicate decoyID rows:", sum(duplicated(bed$decoyID)), "\n")
```

## Step 1: Overlap slop24 with PSI table (HsaIN only, no -Q columns)

```{r load-psi}
psi <- fread(psi_table, sep = "\t", header = TRUE)
psi <- psi[grepl("^HsaIN", EVENT)]
q_cols <- grep("-Q$", names(psi), value = TRUE)
if (length(q_cols) > 0) psi[, (q_cols) := NULL]

coord_parts <- tstrsplit(psi$COORD, ":", fixed = TRUE)
range_parts <- tstrsplit(coord_parts[[2]], "-", fixed = TRUE)
psi[, chr := coord_parts[[1]]]
psi[, intron_start := as.integer(range_parts[[1]])]
psi[, intron_end := as.integer(range_parts[[2]])]
psi <- psi[!is.na(chr) & !is.na(intron_start) & !is.na(intron_end)]
cat("PSI rows retained:", nrow(psi), "\n")
```

```{r overlap}
decoy_gr <- GRanges(seqnames = bed$chr, ranges = IRanges(start = bed$start + 1, end = bed$end), strand = bed$strand)
psi_gr <- GRanges(seqnames = psi$chr, ranges = IRanges(start = psi$intron_start, end = psi$intron_end))
hits <- findOverlaps(decoy_gr, psi_gr, ignore.strand = TRUE)
cat("Overlap pairs:", length(hits), "\n")

bed_hit <- bed[queryHits(hits)]
psi_hit <- psi[subjectHits(hits)]
overlap_cols <- intersect(names(bed_hit), names(psi_hit))
if (length(overlap_cols) > 0) setnames(psi_hit, overlap_cols, paste0("psi_", overlap_cols))
merged <- cbind(bed_hit, psi_hit)

merged[, intron_length := intron_end - intron_start + 1L]
setorder(merged, decoyID, intron_length)
merged <- merged[!duplicated(decoyID)]
cat("Rows after shortest-intron selection:", nrow(merged), "\n")
```

## Step 2: Strand-aware distance from canonical 5'SS

```{r distance}
merged[, decoy_coord_1based := start + 1L]
merged[, decoy_distance_from_canonical_5ss := ifelse(
  strand == "+", decoy_coord_1based - intron_start, intron_end - decoy_coord_1based
)]
```

## Step 3: Inline MaxEntScan (decoy + canonical)

```{r maxent-inline}
read_fasta_sequences <- function(fasta_file) {
  lines <- readLines(fasta_file, warn = FALSE)
  headers <- which(startsWith(lines, ">"))
  if (length(headers) == 0) return(character(0))
  seqs <- character(length(headers))
  for (i in seq_along(headers)) {
    start <- headers[i] + 1
    end <- if (i < length(headers)) headers[i + 1] - 1 else length(lines)
    if (start <= end) {
      seqs[i] <- paste0(lines[start:end], collapse = "")
    } else {
      seqs[i] <- ""
    }
  }
  toupper(seqs)
}

run_maxent_5ss <- function(input_bed, output_bed, genome_fa, bedtools_bin, mode = c("decoy", "canonical")) {
  mode <- match.arg(mode)
  x <- fread(input_bed, sep = "\t", header = FALSE)
  stopifnot(ncol(x) >= 6)
  setnames(x, 1:6, c("chr", "start", "end", "name", "score", "strand"))

  y <- copy(x)
  if (mode == "decoy") {
    y[strand == "+", `:=`(start = pmax(0L, start - 2L), end = end + 6L)]
    y[strand == "-", `:=`(start = pmax(0L, start - 6L), end = end + 2L)]
  } else {
    y[strand == "+", `:=`(start = pmax(0L, start - 4L), end = start + 9L)]
    y[strand == "-", `:=`(start = pmax(0L, end - 6L), end = end + 3L)]
  }

  tmp_bed <- tempfile(fileext = ".bed")
  tmp_fa <- tempfile(fileext = ".fa")
  on.exit(unlink(c(tmp_bed, tmp_fa), force = TRUE), add = TRUE)

  fwrite(y[, c("chr", "start", "end", "name", "score", "strand"), with = FALSE], tmp_bed, sep = "\t", col.names = FALSE)
  system2(bedtools_bin, c("getfasta", "-fi", genome_fa, "-bed", tmp_bed, "-s", "-name", "-fo", tmp_fa), stdout = TRUE, stderr = TRUE)

  seqs <- read_fasta_sequences(tmp_fa)
  seqs <- gsub("N", "A", seqs, fixed = TRUE) # keep sequences valid for MaxEntScan scoring
  scores <- as.numeric(VarCon::calculateMaxEntScanScore(seqs, ssType = 5))
  x[, "maxent_score" := scores]
  fwrite(x, output_bed, sep = "\t", col.names = FALSE)
}

tmp_decoy_bed <- file.path(out_dir, "decoy_introns_for_maxent_decoy_coords.bed")
tmp_canonical_bed <- file.path(out_dir, "decoy_introns_for_maxent_canonical_coords.bed")
maxent_decoy_out <- file.path(out_dir, "decoy_introns_for_maxent_maxent_decoy.bed")
maxent_canonical_out <- file.path(out_dir, "decoy_introns_for_maxent_maxent_canonical.bed")

fwrite(merged[, .(chr, start, end, decoyID, SpliceAI_score, strand)], tmp_decoy_bed, sep = "\t", col.names = FALSE)
canonical_bed <- merged[, .(
  chr = chr,
  start = ifelse(strand == "+", intron_start - 1L, intron_end - 1L),
  end = ifelse(strand == "+", intron_start, intron_end),
  decoyID = decoyID,
  SpliceAI_score = SpliceAI_score,
  strand = strand
)]
fwrite(canonical_bed, tmp_canonical_bed, sep = "\t", col.names = FALSE)

run_maxent_5ss(tmp_decoy_bed, maxent_decoy_out, genome_fa, bedtools_bin, mode = "decoy")
run_maxent_5ss(tmp_canonical_bed, maxent_canonical_out, genome_fa, bedtools_bin, mode = "canonical")
```

## Step 4: Final feature table assembly

```{r merge-features}
decoy_scores_raw <- fread(maxent_decoy_out, sep = "\t", header = FALSE)
canonical_scores_raw <- fread(maxent_canonical_out, sep = "\t", header = FALSE)

decoy_scores <- data.table(decoyID = decoy_scores_raw[[4]], MaxEnt_decoy = as.numeric(decoy_scores_raw[[ncol(decoy_scores_raw)]]))
canonical_scores <- data.table(decoyID = canonical_scores_raw[[4]], MaxEnt_canonical = as.numeric(canonical_scores_raw[[ncol(canonical_scores_raw)]]))

merged <- merge(merged, decoy_scores, by = "decoyID", all.x = TRUE)
merged <- merge(merged, canonical_scores, by = "decoyID", all.x = TRUE)
```

## Step 5: PSI>=10 count, column order, and plot

```{r psi-count}
meta_cols <- c("GENE", "EVENT", "COORD", "LENGTH", "FullCO", "COMPLEX",
               "chr", "intron_start", "intron_end", "intron_length")
feature_cols <- c("decoy_coord_1based", "decoy_distance_from_canonical_5ss",
                  "start", "end", "decoyID", "SpliceAI_score", "strand",
                  "MaxEnt_decoy", "MaxEnt_canonical", "phastCons470_mean", "phastCons100_mean")
psi_cols <- setdiff(names(merged), c(meta_cols, feature_cols, "chr.1"))
psi_num_cols <- psi_cols[sapply(merged[, ..psi_cols], is.numeric)]
merged[, psi_count_ge10 := rowSums(.SD >= 10, na.rm = TRUE), .SDcols = psi_num_cols]
```

```{r reorder-columns}
bed_cols <- c("chr", "start", "end", "decoyID", "SpliceAI_score", "strand")
feature_cols_preferred <- c(
  "decoy_coord_1based", "decoy_distance_from_canonical_5ss",
  "MaxEnt_decoy", "MaxEnt_canonical",
  "phastCons470_mean", "phastCons100_mean", "psi_count_ge10"
)
feature_cols <- feature_cols_preferred[feature_cols_preferred %in% names(merged)]
remaining_cols <- setdiff(names(merged), c(bed_cols, feature_cols))
setcolorder(merged, c(bed_cols, feature_cols, remaining_cols))
```

```{r plot}
plot_df <- merged[!is.na(decoy_distance_from_canonical_5ss) & !is.na(psi_count_ge10)]
plot_df[, log_distance_from_5ss := log10(abs(decoy_distance_from_canonical_5ss) + 1)]
highlight_id <- "HNRNPH1_179620582"

p <- ggplot(
  plot_df,
  aes(
    x = log_distance_from_5ss,
    y = psi_count_ge10,
    size = SpliceAI_score,
    color = phastCons470_mean
  )
) +
  geom_point(alpha = 0.5) +
  geom_point(data = plot_df[decoyID == highlight_id], color = "red", alpha = 1, stroke = 1.2) +
  geom_text(
    data = plot_df[decoyID == highlight_id],
    aes(label = decoyID),
    color = "red", vjust = -0.8, size = 3, show.legend = FALSE
  ) +
  labs(
    title = "Log Distance from Canonical 5'SS vs PSI-count >= 10",
    x = "log10(|decoy distance from canonical 5'SS| + 1)",
    y = "Number of PSI columns >= 10",
    size = "SpliceAI score",
    color = "phastCons 470-way"
  ) +
  scale_color_viridis_c(option = "C", na.value = "grey80") +
  theme_bw()

print(p)
```

## Write Output

```{r write-output}
out_tsv <- file.path(out_dir, "decoy_intron_features_draft_v2_inline.tsv")
fwrite(merged, out_tsv, sep = "\t")
cat("Output written:", out_tsv, "\n")
cat("Rows:", nrow(merged), " Columns:", ncol(merged), "\n")
```
