---
title: "Conservation and Intron Retention Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-libraries}
library(data.table)
library(GenomicRanges)
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)
library(gridExtra)
```

## Load Data

```{r load-data}
# Load control SpliceAI data
control_spliceai <- fread("control_spliceai_filtered_protein_coding.tsv")

# Load SpliceAI MAF data and add _D suffix to column names
spliceai_maf <- fread("Top50_Decoys_Clusters2_4_MaxSpliceAI.tsv")
colnames(spliceai_maf) <- paste0(colnames(spliceai_maf), "_D")

# Load PSI table
PSI_table <- fread("PSI_TABLE-hg38.tab.gz")

# Load phyloP decoy data and add _D suffix to column names
phylop_decoy <- fread("phyloP/decoy/phyloP_decoy_protein_coding.tsv")
colnames(phylop_decoy) <- paste0(colnames(phylop_decoy), "_D")

# Load phyloP control data
phylop_ctrl <- fread("phyloP/ctrl/phyloP_ctrl_summary_filtered_protein_coding.tsv")
```

## Extract Decoy Coordinates and Create Extended BED File

```{r extract-decoy-coords}
# Extract decoyIDs from the filtered SpliceAI file
# Note: column names have _D suffix, so decoyID is now decoyID_D
# Also, the decoyID values themselves have _D suffix (e.g., "ACTB_5530194_D")
# But decoyIDs.bed has IDs without _D suffix (e.g., "ACTB_5530194")
# So we need to remove the _D suffix from the values for matching
decoy_ids_filtered_raw <- spliceai_maf$decoyID_D
decoy_ids_filtered <- gsub("_D$", "", decoy_ids_filtered_raw)  # Remove _D suffix

cat("Number of filtered decoys:", length(decoy_ids_filtered), "\n")
cat("Sample decoyIDs (first 5):", paste(head(decoy_ids_filtered, 5), collapse = ", "), "\n")

# Load decoyIDs.bed to get coordinates
decoy_coords_full <- fread("decoyIDs.bed", 
                           col.names = c("chr", "start", "end", "decoyID", "score", "strand", "other"))

# Match filtered decoys with their coordinates
decoy_coords_matched <- decoy_coords_full[decoyID %in% decoy_ids_filtered, ]

cat("Decoys with coordinates found:", nrow(decoy_coords_matched), "\n")
cat("Decoys missing coordinates:", sum(!decoy_ids_filtered %in% decoy_coords_full$decoyID), "\n")

# Get hg38 SpliceAI scores (human scores) from the filtered file
# The hg38 column should be named hg38_D after adding _D suffix
hg38_col <- grep("^hg38", colnames(spliceai_maf), value = TRUE)
if (length(hg38_col) == 0) {
  stop("Could not find hg38 column in spliceai_maf. Available columns:", 
       paste(head(colnames(spliceai_maf), 10), collapse = ", "))
}
if (length(hg38_col) > 1) {
  # Take the first match
  hg38_col <- hg38_col[1]
  cat("Multiple hg38 columns found, using:", hg38_col, "\n")
}
cat("Using hg38 column:", hg38_col, "\n")

# Create a data frame with decoyID (without _D suffix) and hg38 score
decoy_scores <- data.frame(
  decoyID = decoy_ids_filtered,  # Use the cleaned IDs without _D suffix
  decoyID_original = decoy_ids_filtered_raw,  # Keep original for reference
  hg38_score = as.numeric(spliceai_maf[[hg38_col]])
)

# Merge coordinates with scores
decoy_coords_with_scores <- merge(
  decoy_coords_matched[, .(decoyID, chr, start, end, strand)],
  decoy_scores,
  by = "decoyID",
  all.x = TRUE
)

cat("Decoys with both coordinates and scores:", sum(!is.na(decoy_coords_with_scores$hg38_score)), "\n")
```

## Overlap Decoy Sites with Intron Coordinates

```{r overlap-decoys-introns}
# Create GRanges for decoy sites (before coordinate adjustment)
decoy_gr_sites <- GRanges(
  seqnames = decoy_coords_with_scores$chr,
  ranges = IRanges(start = decoy_coords_with_scores$start, 
                   end = decoy_coords_with_scores$end),
  decoyID = decoy_coords_with_scores$decoyID,
  strand = decoy_coords_with_scores$strand,
  hg38_score = decoy_coords_with_scores$hg38_score
)

# We need intron coordinates from PSI table - but we haven't filtered it yet
# So let's create a temporary intron GRanges from the full PSI table for overlap
# Parse COORD column to get intron boundaries
parse_coord_temp <- function(coord_str) {
  parts <- strsplit(coord_str, ":")[[1]]
  if (length(parts) == 2) {
    chr <- parts[1]
    coords <- strsplit(parts[2], "-")[[1]]
    if (length(coords) == 2) {
      start <- as.numeric(coords[1])
      end <- as.numeric(coords[2])
      return(data.frame(chr = chr, start = start, end = end))
    }
  }
  return(data.frame(chr = NA, start = NA, end = NA))
}

# Load full PSI table temporarily for intron coordinates
PSI_table_temp <- fread("PSI_TABLE-hg38.tab.gz")
PSI_table_temp <- PSI_table_temp[grepl("^HsaIN", EVENT), ]

# Extract intron coordinates
intron_coords_temp <- do.call(rbind, lapply(PSI_table_temp$COORD, parse_coord_temp))
intron_coords_temp$EVENT <- PSI_table_temp$EVENT

# Create GRanges for introns
intron_gr_temp <- GRanges(
  seqnames = intron_coords_temp$chr,
  ranges = IRanges(start = intron_coords_temp$start, end = intron_coords_temp$end),
  EVENT = intron_coords_temp$EVENT
)

# Find overlaps between decoy sites and introns
overlaps_decoy_intron <- findOverlaps(decoy_gr_sites, intron_gr_temp, ignore.strand = TRUE)

# Get the full intron coordinates for each decoy
# Note: decoy_gr_sites has decoyID without _D suffix (from decoy_coords_with_scores)
decoy_intron_matched <- data.frame(
  decoyID = decoy_gr_sites$decoyID[queryHits(overlaps_decoy_intron)],
  decoy_chr = seqnames(decoy_gr_sites)[queryHits(overlaps_decoy_intron)],
  decoy_start = start(decoy_gr_sites)[queryHits(overlaps_decoy_intron)],
  decoy_end = end(decoy_gr_sites)[queryHits(overlaps_decoy_intron)],
  decoy_strand = strand(decoy_gr_sites)[queryHits(overlaps_decoy_intron)],
  hg38_score = decoy_gr_sites$hg38_score[queryHits(overlaps_decoy_intron)],
  intron_chr = seqnames(intron_gr_temp)[subjectHits(overlaps_decoy_intron)],
  intron_start = start(intron_gr_temp)[subjectHits(overlaps_decoy_intron)],
  intron_end = end(intron_gr_temp)[subjectHits(overlaps_decoy_intron)],
  EVENT = intron_gr_temp$EVENT[subjectHits(overlaps_decoy_intron)]
)

# Calculate intron length
decoy_intron_matched$intron_length <- decoy_intron_matched$intron_end - decoy_intron_matched$intron_start

cat("Decoy sites overlapping introns (before filtering):", nrow(decoy_intron_matched), "\n")
cat("Unique decoys with intron overlaps (before filtering):", length(unique(decoy_intron_matched$decoyID)), "\n")

# If a decoy overlaps multiple introns, select the one with the shortest length
decoy_intron_matched <- decoy_intron_matched %>%
  arrange(decoyID, intron_length) %>%  # Sort by decoyID, then by intron length (shortest first)
  distinct(decoyID, .keep_all = TRUE)  # Keep only the first match (shortest intron) for each decoy

cat("Decoy sites overlapping introns (after selecting shortest):", nrow(decoy_intron_matched), "\n")
cat("Unique decoys with intron overlaps (after filtering):", length(unique(decoy_intron_matched$decoyID)), "\n")

# Clean up temporary objects
rm(PSI_table_temp, intron_coords_temp, intron_gr_temp)
gc()
```

## Adjust Coordinates and Create Final BED File

```{r create-extended-bed}
# Use the full intron coordinates (intron_start, intron_end)
# Adjust the start coordinate based on strand to include canonical 5' splice sites

bed_data <- decoy_intron_matched %>%
  mutate(
    # Use intron coordinates as the base
    chr = as.character(intron_chr),
    start = intron_start,
    end = intron_end,
    name = decoyID,
    score = hg38_score,
    strand = as.character(decoy_strand)
  ) %>%
  # Adjust coordinates based on strand to include canonical 5' splice sites
  # For positive strand (+): subtract 50 from start (extend upstream to include 5' splice site)
  # For negative strand (-): add 50 to end (extend upstream in genomic coords, which is the 5' splice site direction)
  mutate(
    start_adjusted = ifelse(strand == "+", 
                           start - 50,  # Subtract 50 for positive strand (move upstream)
                           start),      # Keep start unchanged for negative strand
    end_adjusted = ifelse(strand == "+",
                         end,           # Keep end unchanged for positive strand
                         end + 50)      # Add 50 to end for negative strand (extend upstream)
  ) %>%
  # Ensure start is not negative
  mutate(start_adjusted = pmax(1, start_adjusted)) %>%
  # Ensure start < end
  mutate(start_adjusted = pmin(start_adjusted, end_adjusted - 1)) %>%
  # Select and rename columns for BED format
  # Use decoyID (without _D) as the name in BED file
  select(chr, start = start_adjusted, end = end_adjusted, name = decoyID, score = hg38_score, strand) %>%
  # Remove rows with missing data
  filter(!is.na(chr) & !is.na(start) & !is.na(end) & !is.na(score)) %>%
  distinct(name, .keep_all = TRUE)

cat("Final BED entries:", nrow(bed_data), "\n")
cat("Range of coordinates:\n")
cat("  Start range:", min(bed_data$start), "to", max(bed_data$start), "\n")
cat("  End range:", min(bed_data$end), "to", max(bed_data$end), "\n")
cat("  Strand distribution:\n")
print(table(bed_data$strand))

# Validate BED data
bed_data <- bed_data %>%
  # Ensure start < end
  filter(start < end) %>%
  # Sort by chromosome and start position
  arrange(chr, start)

cat("Validated BED entries:", nrow(bed_data), "\n")

# Write BED file (6-column format: chr, start, end, name, score, strand)
bed_output <- "DecoysInConservedGenes50thpercentilev2.bed"
fwrite(bed_data, 
       file = bed_output,
       sep = "\t",
       col.names = FALSE,
       quote = FALSE)

cat("\nBED file written to:", bed_output, "\n")
cat("File contains", nrow(bed_data), "entries\n")
cat("BED format: chr, start, end, decoyID, hg38_SpliceAI_score, strand\n")

# Display first few lines
cat("\nFirst 5 entries:\n")
print(head(bed_data, 5))

# Summary statistics
cat("\n=== BED File Summary ===\n")
cat("Chromosomes:", length(unique(bed_data$chr)), "\n")
cat("Score range:", min(bed_data$score, na.rm = TRUE), "to", max(bed_data$score, na.rm = TRUE), "\n")
cat("Coordinate length range:", min(bed_data$end - bed_data$start), "to", max(bed_data$end - bed_data$start), "bp\n")
```
```

## Filter PSI Table

```{r filter-psi}
# Filter to only rows where EVENT starts with "HsaIN" (matching reference pattern)
PSI_table <- PSI_table[grepl("^HsaIN", EVENT), ]

cat("Number of HsaIN events:", nrow(PSI_table), "\n")

# Remove columns ending in "-Q"
q_cols <- grep("-Q$", colnames(PSI_table), value = TRUE)
if (length(q_cols) > 0) {
  PSI_table[, (q_cols) := NULL]
  cat("Removed", length(q_cols), "quality columns\n")
}

# Identify PSI value columns (exclude metadata columns)
metadata_cols <- c("GENE", "EVENT", "COORD", "LENGTH", "FullCO", "COMPLEX")
psi_cols <- setdiff(colnames(PSI_table), metadata_cols)

cat("Number of PSI value columns:", length(psi_cols), "\n")

# Remove rows where >80% of PSI values are NA
na_threshold <- 0.8
if (length(psi_cols) > 0 && nrow(PSI_table) > 0) {
  row_na_prop <- rowSums(is.na(PSI_table[, ..psi_cols])) / length(psi_cols)
  PSI_table <- PSI_table[row_na_prop <= na_threshold, ]
  cat("Rows after filtering >80% NA:", nrow(PSI_table), "\n")
} else {
  stop("No PSI value columns found or table is empty")
}

# Remove columns where >80% of PSI values are NA
if (nrow(PSI_table) > 0 && length(psi_cols) > 0) {
  col_na_prop <- colSums(is.na(PSI_table[, ..psi_cols])) / nrow(PSI_table)
  cols_to_keep <- names(col_na_prop[col_na_prop <= na_threshold])
  psi_cols_filtered <- c(metadata_cols, cols_to_keep)
  PSI_table <- PSI_table[, ..psi_cols_filtered]
  cat("Columns after filtering >80% NA:", ncol(PSI_table), "\n")
}

# Check final table size
if (nrow(PSI_table) == 0) {
  stop("PSI table is empty after filtering. Check filtering criteria.")
}

# Create filtered version and remove original to free up space
PSI_table_filtered <- PSI_table
rm(PSI_table)
gc()

cat("Final PSI table dimensions:", nrow(PSI_table_filtered), "rows x", ncol(PSI_table_filtered), "columns\n")
```

## Combine phyloP Tables

```{r combine-phylop}
# Combine the two phyloP tables
# First, ensure column names match (phylop_ctrl has decoyID, phylop_decoy has MAF_file and decoy_id_clean_D)
# We'll use decoyID as the common identifier

# Rename columns in phylop_decoy to match structure
# phylop_decoy has: MAF_file_D, mean_phylop_D, max_phylop_D, etc., decoy_id_clean_D
# phylop_ctrl has: decoyID, mean_phylop, max_phylop, etc.

# Create a unified phyloP table
phylop_combined <- rbind(
  phylop_ctrl %>% 
    mutate(source = "control") %>%
    select(decoyID, mean_phylop, max_phylop, fraction_constrained, 
           fraction_constrained_p01, sum_positive_phylop, source),
  phylop_decoy %>% 
    mutate(source = "decoy") %>%
    rename(decoyID = decoy_id_clean_D) %>%
    select(decoyID, mean_phylop_D, max_phylop_D, fraction_constrained_D, 
           fraction_constrained_p01_D, sum_positive_phylop_D, source) %>%
    rename(mean_phylop = mean_phylop_D,
           max_phylop = max_phylop_D,
           fraction_constrained = fraction_constrained_D,
           fraction_constrained_p01 = fraction_constrained_p01_D,
           sum_positive_phylop = sum_positive_phylop_D)
)
```

## Extract Intron Coordinates from PSI Table

```{r extract-intron-coords}
# Parse COORD column to extract chromosome, start, and end
# COORD format appears to be like "chrX:100636191-100636207"
parse_coord <- function(coord_str) {
  # Extract chromosome and coordinates
  parts <- strsplit(coord_str, ":")[[1]]
  if (length(parts) == 2) {
    chr <- parts[1]
    coords <- strsplit(parts[2], "-")[[1]]
    if (length(coords) == 2) {
      start <- as.numeric(coords[1])
      end <- as.numeric(coords[2])
      return(data.frame(chr = chr, start = start, end = end))
    }
  }
  return(data.frame(chr = NA, start = NA, end = NA))
}

# Apply parsing to all rows
intron_coords <- do.call(rbind, lapply(PSI_table_filtered$COORD, parse_coord))
PSI_table_filtered <- cbind(PSI_table_filtered, intron_coords)

# Create GRanges object for introns
intron_gr <- GRanges(
  seqnames = PSI_table_filtered$chr,
  ranges = IRanges(start = PSI_table_filtered$start, end = PSI_table_filtered$end),
  EVENT = PSI_table_filtered$EVENT,
  GENE = PSI_table_filtered$GENE
)
```

## Load decoyID to Coordinate Mapping

```{r load-decoyid-mapping}
# Load decoyIDs.bed file which maps decoyID to genomic coordinates
# Format: chr, start, end, decoyID, score, strand, other
decoy_coords <- fread("decoyIDs.bed", 
                      col.names = c("chr", "start", "end", "decoyID", "score", "strand", "other"))

# Load control BED file which maps control IDs to genomic coordinates
# Format: chr, start, end, ID (column 4), score, strand
control_coords <- fread("MatchedControlSpliceSites_uniqueIDsorted.bed",
                        col.names = c("chr", "start", "end", "ID", "score", "strand"))

# Create a mapping for control data
# Control decoyIDs have _C or _P suffixes (e.g., "GENE_COORD_C" or "GENE_COORD_P")
# The BED file column 4 has IDs without suffixes (e.g., "GENE_COORD")
# We need to remove the suffix to match
# Create a lookup: base ID -> coordinates
control_coords_lookup <- control_coords[, .(chr, start, end, strand, ID)]
setnames(control_coords_lookup, "ID", "base_id")

# Merge decoy coordinates (direct match on decoyID)
decoy_coords_merge <- decoy_coords[, .(decoyID, chr, start, end, strand)]
decoy_phylop_coords <- merge(
  phylop_combined[source == "decoy", ],
  decoy_coords_merge,
  by = "decoyID",
  all.x = TRUE
)

# For control decoyIDs, extract the base ID (remove _C or _P suffix) to match BED file column 4
phylop_control <- phylop_combined[source == "control", ]
phylop_control$base_id <- gsub("_(C|P)$", "", phylop_control$decoyID)

# Merge control coordinates (match on base_id)
control_phylop_coords <- merge(
  phylop_control,
  control_coords_lookup,
  by = "base_id",
  all.x = TRUE
)

# Combine decoy and control coordinates
# Select only the needed columns, excluding base_id
phylop_with_coords <- rbind(
  decoy_phylop_coords[, .(decoyID, chr, start, end, strand, mean_phylop, max_phylop, 
                          fraction_constrained, fraction_constrained_p01, sum_positive_phylop, source)],
  control_phylop_coords[, .(decoyID, chr, start, end, strand, mean_phylop, max_phylop,
                            fraction_constrained, fraction_constrained_p01, sum_positive_phylop, source)]
)

cat("Coordinates merged:\n")
cat("  Decoy sites with coordinates:", sum(!is.na(decoy_phylop_coords$chr)), "/", nrow(decoy_phylop_coords), "\n")
cat("  Control sites with coordinates:", sum(!is.na(control_phylop_coords$chr)), "/", nrow(control_phylop_coords), "\n")
cat("  Total sites with coordinates:", sum(!is.na(phylop_with_coords$chr)), "/", nrow(phylop_with_coords), "\n")

# Create GRanges object for decoy sites (for reference)
decoy_gr <- GRanges(
  seqnames = decoy_coords$chr,
  ranges = IRanges(start = decoy_coords$start, end = decoy_coords$end),
  decoyID = decoy_coords$decoyID,
  strand = decoy_coords$strand
)

# Create GRanges object for phyloP sites
phylop_gr <- GRanges(
  seqnames = phylop_with_coords$chr,
  ranges = IRanges(start = phylop_with_coords$start, end = phylop_with_coords$end),
  decoyID = phylop_with_coords$decoyID,
  sum_positive_phylop = phylop_with_coords$sum_positive_phylop,
  mean_phylop = phylop_with_coords$mean_phylop,
  source = phylop_with_coords$source
)
```

## Calculate Mean Intron Retention per Intron

```{r calculate-mean-psi}
# Calculate mean PSI (intron retention) across all samples for each intron
psi_cols_final <- setdiff(colnames(PSI_table_filtered), 
                         c("GENE", "EVENT", "COORD", "LENGTH", "FullCO", "COMPLEX", 
                           "chr", "start", "end"))

PSI_table_filtered$mean_PSI <- rowMeans(PSI_table_filtered[, ..psi_cols_final], na.rm = TRUE)
PSI_table_filtered$median_PSI <- apply(PSI_table_filtered[, ..psi_cols_final], 1, median, na.rm = TRUE)
```

## Overlap phyloP Sites with Introns

```{r overlap-phylop-introns}
# Find overlaps between phyloP sites and introns
# Using findOverlaps to match phyloP sites that fall within intron boundaries
overlaps <- findOverlaps(phylop_gr, intron_gr, ignore.strand = TRUE)

# Create a data frame with matched phyloP and intron retention data
matched_data <- data.frame(
  decoyID = phylop_gr$decoyID[queryHits(overlaps)],
  sum_positive_phylop = phylop_gr$sum_positive_phylop[queryHits(overlaps)],
  mean_phylop = phylop_gr$mean_phylop[queryHits(overlaps)],
  source = phylop_gr$source[queryHits(overlaps)],
  EVENT = intron_gr$EVENT[subjectHits(overlaps)],
  GENE = intron_gr$GENE[subjectHits(overlaps)],
  intron_idx = subjectHits(overlaps)
)

# Add intron retention values
matched_data$mean_PSI <- PSI_table_filtered$mean_PSI[matched_data$intron_idx]
matched_data$median_PSI <- PSI_table_filtered$median_PSI[matched_data$intron_idx]

# Calculate intron length from PSI table
matched_data$intron_length <- PSI_table_filtered$LENGTH[matched_data$intron_idx]

cat("Number of phyloP sites overlapping introns (before filtering):", nrow(matched_data), "\n")

# If a decoy overlaps multiple introns, select the one with the shortest length
matched_data <- matched_data %>%
  arrange(decoyID, intron_length) %>%  # Sort by decoyID, then by intron length (shortest first)
  distinct(decoyID, .keep_all = TRUE)  # Keep only the first match (shortest intron) for each decoy

# Remove rows with missing data
matched_data <- matched_data[complete.cases(matched_data[, c("sum_positive_phylop", "mean_PSI")]), ]

cat("Number of phyloP sites overlapping introns (after selecting shortest):", nrow(matched_data), "\n")
cat("Number of unique introns with phyloP data:", length(unique(matched_data$intron_idx)), "\n")
```

## Correlation Analysis: phyloP vs Intron Retention

```{r correlation-analysis}
# Identify site type based on decoyID suffix
# _C = canonical control, _P = nonPRPF8 control, otherwise = decoy
matched_data$site_type <- ifelse(grepl("_C$", matched_data$decoyID), "Canonical (_C)",
                                  ifelse(grepl("_P$", matched_data$decoyID), "nonPRPF8 (_P)", 
                                         "Decoy"))
matched_data$site_type <- factor(matched_data$site_type, 
                                  levels = c("Decoy", "Canonical (_C)", "nonPRPF8 (_P)"))

# Calculate correlation for each type separately
cor_decoy <- cor(matched_data$sum_positive_phylop[matched_data$site_type == "Decoy"], 
                 matched_data$mean_PSI[matched_data$site_type == "Decoy"], 
                 use = "complete.obs")
cor_canonical <- cor(matched_data$sum_positive_phylop[matched_data$site_type == "Canonical (_C)"], 
                     matched_data$mean_PSI[matched_data$site_type == "Canonical (_C)"], 
                     use = "complete.obs")
cor_nonprpf8 <- cor(matched_data$sum_positive_phylop[matched_data$site_type == "nonPRPF8 (_P)"], 
                    matched_data$mean_PSI[matched_data$site_type == "nonPRPF8 (_P)"], 
                    use = "complete.obs")

# Overall correlation
cor_overall <- cor(matched_data$sum_positive_phylop, matched_data$mean_PSI, use = "complete.obs")

cat("=== Correlation: Sum Positive phyloP vs Intron Retention ===\n")
cat("Overall correlation:", round(cor_overall, 4), "\n")
cat("Decoy correlation:", round(cor_decoy, 4), "\n")
cat("Canonical (_C) correlation:", round(cor_canonical, 4), "\n")
cat("nonPRPF8 (_P) correlation:", round(cor_nonprpf8, 4), "\n")

# Count sites by type
cat("\n=== Site Counts by Type ===\n")
print(table(matched_data$site_type))

# Plot correlation with three colors
library(ggplot2)
p1 <- ggplot(matched_data, aes(x = sum_positive_phylop, y = mean_PSI, color = site_type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, aes(fill = site_type)) +
  scale_color_manual(values = c("Decoy" = "#E31A1C", 
                                 "Canonical (_C)" = "#1F78B4", 
                                 "nonPRPF8 (_P)" = "#33A02C"),
                      name = "Site Type") +
  scale_fill_manual(values = c("Decoy" = "#E31A1C", 
                                "Canonical (_C)" = "#1F78B4", 
                                "nonPRPF8 (_P)" = "#33A02C"),
                     name = "Site Type") +
  labs(title = "phyloP Conservation vs Intron Retention by Site Type",
       subtitle = paste0("Decoy r=", round(cor_decoy, 3), 
                        " | Canonical r=", round(cor_canonical, 3),
                        " | nonPRPF8 r=", round(cor_nonprpf8, 3)),
       x = "Sum Positive phyloP",
       y = "Mean PSI (Intron Retention)") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p1)

# Create separate plots for each type for clarity
p2 <- ggplot(matched_data, aes(x = sum_positive_phylop, y = mean_PSI)) +
  geom_point(alpha = 0.3, color = "#E31A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E31A1C", fill = "#E31A1C") +
  labs(title = "Decoy: phyloP Conservation vs Intron Retention",
       subtitle = paste0("Correlation = ", round(cor_decoy, 4)),
       x = "Sum Positive phyloP",
       y = "Mean PSI (Intron Retention)") +
  theme_minimal()

p3 <- ggplot(matched_data[matched_data$site_type == "Canonical (_C)", ], 
             aes(x = sum_positive_phylop, y = mean_PSI)) +
  geom_point(alpha = 0.3, color = "#1F78B4") +
  geom_smooth(method = "lm", se = TRUE, color = "#1F78B4", fill = "#1F78B4") +
  labs(title = "Canonical (_C): phyloP Conservation vs Intron Retention",
       subtitle = paste0("Correlation = ", round(cor_canonical, 4)),
       x = "Sum Positive phyloP",
       y = "Mean PSI (Intron Retention)") +
  theme_minimal()

p4 <- ggplot(matched_data[matched_data$site_type == "nonPRPF8 (_P)", ], 
             aes(x = sum_positive_phylop, y = mean_PSI)) +
  geom_point(alpha = 0.3, color = "#33A02C") +
  geom_smooth(method = "lm", se = TRUE, color = "#33A02C", fill = "#33A02C") +
  labs(title = "nonPRPF8 (_P): phyloP Conservation vs Intron Retention",
       subtitle = paste0("Correlation = ", round(cor_nonprpf8, 4)),
       x = "Sum Positive phyloP",
       y = "Mean PSI (Intron Retention)") +
  theme_minimal()

# Print individual plots
print(p2)
print(p3)
print(p4)
```

## Prepare Data for Logistic Regression

```{r prepare-logistic-data}
# For logistic regression, we need to:
# 1. Aggregate phyloP scores per intron (since multiple phyloP sites may overlap one intron)
# 2. Create a binary outcome variable for intron retention (high vs low)
# 3. Use conservation (phyloP) as predictor

# Aggregate phyloP scores per intron (take mean or sum of overlapping sites)
intron_phylop <- matched_data %>%
  group_by(intron_idx) %>%
  summarise(
    mean_sum_positive_phylop = mean(sum_positive_phylop, na.rm = TRUE),
    sum_positive_phylop_total = sum(sum_positive_phylop, na.rm = TRUE),
    mean_mean_phylop = mean(mean_phylop, na.rm = TRUE),
    n_phylop_sites = n(),
    source = paste(unique(source), collapse = ","),
    .groups = "drop"
  )

# Merge with intron retention data
regression_data <- PSI_table_filtered %>%
  mutate(intron_idx = row_number()) %>%
  left_join(intron_phylop, by = "intron_idx") %>%
  filter(!is.na(mean_sum_positive_phylop))  # Only keep introns with phyloP data

# Create binary outcome variable
# Classify introns as "retained" (1) if mean_PSI > median, else "not retained" (0)
median_psi_threshold <- median(regression_data$mean_PSI, na.rm = TRUE)
regression_data$retained <- ifelse(regression_data$mean_PSI > median_psi_threshold, 1, 0)

cat("Median PSI threshold:", round(median_psi_threshold, 4), "\n")
cat("Number of retained introns:", sum(regression_data$retained == 1), "\n")
cat("Number of not retained introns:", sum(regression_data$retained == 0), "\n")
cat("Total introns with phyloP data:", nrow(regression_data), "\n")
```

## Logistic Regression Model

```{r logistic-regression}
# Fit logistic regression model
# Model 1: Using sum_positive_phylop as predictor (median threshold)
model1 <- glm(retained ~ mean_sum_positive_phylop, 
              data = regression_data, 
              family = binomial(link = "logit"))

summary(model1)

# Model 2: Using mean_phylop as predictor
model2 <- glm(retained ~ mean_mean_phylop, 
              data = regression_data, 
              family = binomial(link = "logit"))

summary(model2)

# Model 3: Using both predictors
model3 <- glm(retained ~ mean_sum_positive_phylop + mean_mean_phylop, 
              data = regression_data, 
              family = binomial(link = "logit"))

summary(model3)

# Model 4: Include source as a factor
if (length(unique(regression_data$source)) > 1) {
  regression_data$source_factor <- as.factor(regression_data$source)
  model4 <- glm(retained ~ mean_sum_positive_phylop + source_factor, 
                data = regression_data, 
                family = binomial(link = "logit"))
  
  summary(model4)
}

# Calculate odds ratios and confidence intervals
library(broom)
model1_summary <- tidy(model1, conf.int = TRUE, exponentiate = TRUE)
print(model1_summary)

# Visualize model predictions
regression_data$predicted_prob <- predict(model1, type = "response")

p3 <- ggplot(regression_data, aes(x = mean_sum_positive_phylop, y = retained)) +
  geom_point(alpha = 0.3, position = position_jitter(height = 0.05)) +
  geom_line(aes(y = predicted_prob), color = "red", size = 1) +
  labs(title = "Logistic Regression: Conservation Predicting Intron Retention (Median Threshold)",
       x = "Mean Sum Positive phyloP",
       y = "Intron Retained (1) or Not (0)") +
  theme_minimal()

print(p3)
```

## Additional Logistic Regression Models with PSI Thresholds

```{r logistic-regression-thresholds}
# Create binary outcome variables with different PSI thresholds
# Threshold 1: >10 PSI = 0 (not retained), <=10 PSI = 1 (retained)
regression_data$retained_10 <- ifelse(regression_data$mean_PSI <= 10, 1, 0)

# Threshold 2: >15 PSI = 0 (not retained), <=15 PSI = 1 (retained)
regression_data$retained_15 <- ifelse(regression_data$mean_PSI <= 15, 1, 0)

cat("PSI Threshold 10:\n")
cat("  Retained (<=10 PSI):", sum(regression_data$retained_10 == 1), "\n")
cat("  Not retained (>10 PSI):", sum(regression_data$retained_10 == 0), "\n")

cat("\nPSI Threshold 15:\n")
cat("  Retained (<=15 PSI):", sum(regression_data$retained_15 == 1), "\n")
cat("  Not retained (>15 PSI):", sum(regression_data$retained_15 == 0), "\n")

# Model 5: PSI threshold 10
model5 <- glm(retained_10 ~ mean_sum_positive_phylop, 
              data = regression_data, 
              family = binomial(link = "logit"))

cat("\n=== Model 5: PSI Threshold <= 10 ===\n")
summary(model5)

model5_summary <- tidy(model5, conf.int = TRUE, exponentiate = TRUE)
print(model5_summary)

# Model 6: PSI threshold 15
model6 <- glm(retained_15 ~ mean_sum_positive_phylop, 
              data = regression_data, 
              family = binomial(link = "logit"))

cat("\n=== Model 6: PSI Threshold <= 15 ===\n")
summary(model6)

model6_summary <- tidy(model6, conf.int = TRUE, exponentiate = TRUE)
print(model6_summary)

# Visualize model predictions for threshold models
regression_data$predicted_prob_10 <- predict(model5, type = "response")
regression_data$predicted_prob_15 <- predict(model6, type = "response")

p4 <- ggplot(regression_data, aes(x = mean_sum_positive_phylop, y = retained_10)) +
  geom_point(alpha = 0.3, position = position_jitter(height = 0.05)) +
  geom_line(aes(y = predicted_prob_10), color = "blue", size = 1) +
  labs(title = "Logistic Regression: Conservation Predicting Intron Retention (PSI <= 10)",
       x = "Mean Sum Positive phyloP",
       y = "Intron Retained (1 if PSI <= 10, 0 if PSI > 10)") +
  theme_minimal()

print(p4)

p5 <- ggplot(regression_data, aes(x = mean_sum_positive_phylop, y = retained_15)) +
  geom_point(alpha = 0.3, position = position_jitter(height = 0.05)) +
  geom_line(aes(y = predicted_prob_15), color = "green", size = 1) +
  labs(title = "Logistic Regression: Conservation Predicting Intron Retention (PSI <= 15)",
       x = "Mean Sum Positive phyloP",
       y = "Intron Retained (1 if PSI <= 15, 0 if PSI > 15)") +
  theme_minimal()

print(p5)

# Compare all three models side by side
library(gridExtra)
p6 <- ggplot(regression_data, aes(x = mean_sum_positive_phylop, y = retained)) +
  geom_point(alpha = 0.2, position = position_jitter(height = 0.05)) +
  geom_line(aes(y = predicted_prob), color = "red", size = 1) +
  labs(title = "Median Threshold",
       x = "Mean Sum Positive phyloP",
       y = "Retained") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

p7 <- ggplot(regression_data, aes(x = mean_sum_positive_phylop, y = retained_10)) +
  geom_point(alpha = 0.2, position = position_jitter(height = 0.05)) +
  geom_line(aes(y = predicted_prob_10), color = "blue", size = 1) +
  labs(title = "PSI <= 10",
       x = "Mean Sum Positive phyloP",
       y = "Retained") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

p8 <- ggplot(regression_data, aes(x = mean_sum_positive_phylop, y = retained_15)) +
  geom_point(alpha = 0.2, position = position_jitter(height = 0.05)) +
  geom_line(aes(y = predicted_prob_15), color = "green", size = 1) +
  labs(title = "PSI <= 15",
       x = "Mean Sum Positive phyloP",
       y = "Retained") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(p6, p7, p8, ncol = 3, 
             top = "Logistic Regression Models: Conservation Predicting Intron Retention")
```

## Data Summary

```{r data-summary}
cat("\n=== Final Data Summary ===\n")
cat("PSI table (intron retention):", nrow(PSI_table_filtered), "introns\n")
cat("phyloP control:", nrow(phylop_ctrl), "sites\n")
cat("phyloP decoy:", nrow(phylop_decoy), "sites\n")
cat("phyloP combined:", nrow(phylop_combined), "sites\n")
cat("Control SpliceAI:", nrow(control_spliceai), "sites\n")
cat("SpliceAI MAF:", nrow(spliceai_maf), "sites\n")
cat("Matched phyloP-intron pairs:", nrow(matched_data), "\n")
cat("Introns with phyloP data for regression:", nrow(regression_data), "\n")
```

## Summary

This analysis has:
1. ✅ Loaded all required data files
2. ✅ Added "_D" suffixes to SpliceAI_MAF and phyloP_decoy columns
3. ✅ Filtered PSI table to HsaIN. events only
4. ✅ Removed "-Q" columns from PSI table
5. ✅ Filtered rows and columns with >80% NA values
6. ✅ Combined phyloP tables
7. ✅ Extracted intron coordinates from PSI table
8. ✅ Loaded decoyID to coordinate mapping from decoyIDs.bed
9. ✅ Overlapped phyloP scores with intron coordinates using GenomicRanges
10. ✅ Performed correlation analysis between phyloP and intron retention
11. ✅ Completed logistic regression models to predict intron retention from conservation

**Key Findings:**
- Correlation coefficients between phyloP scores and intron retention
- Logistic regression models showing how conservation predicts intron retention
- Visualizations of the relationships
