---
title: "SpliceAI MAF Heatmap"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 16, fig.height = 12)
```

## Load Libraries

```{r libraries}
library(data.table)
library(pheatmap)
library(viridis)
library(beepr)
library(gridExtra)
library(grid)
library(dplyr)
```

## Load Data

```{r load-data}
# Read the TSV file (filtered for min 0.01 SpliceAI score in human)
spliceai_data <- fread("SpliceAI_MAF_protein_coding_min0.01.tsv", sep = "\t", header = TRUE)
cat("Decoy dataset loaded (min 0.01 SpliceAI score in human):\n")
cat("  Number of decoy IDs:", nrow(spliceai_data), "\n")
cat("  Number of species:", ncol(spliceai_data) - 1, "\n")
```

## Prepare Data Matrix for K-means Clustering (with -1 values)

```{r prepare-matrix-kmeans}
# Extract decoy IDs (first column)
decoy_ids_kmeans <- spliceai_data$decoyID

# Extract numeric data (all columns except first)
data_matrix_kmeans <- as.matrix(spliceai_data[, -1, with = FALSE])

# Set row names
rownames(data_matrix_kmeans) <- decoy_ids_kmeans

# Read species order and mapping
species_order <- fread("Newickspeciesorder.txt", header = FALSE, col.names = "tree_name")
species_mapping <- fread("tree_to_clade_mapping.tsv", sep = "\t", header = TRUE)
ordered_species <- merge(species_order, species_mapping, by = "tree_name", all.x = TRUE, sort = FALSE)
ordered_species_ids <- ordered_species$species_id[!is.na(ordered_species$species_id)]

# Reorder columns to match Newick order
data_matrix_kmeans <- data_matrix_kmeans[, ordered_species_ids, drop = FALSE]

# Remove rows that are entirely -1 (can't cluster these)
rows_all_neg1 <- rowSums(data_matrix_kmeans != -1) == 0
if (sum(rows_all_neg1) > 0) {
  cat("K-means matrix: Removing", sum(rows_all_neg1), "rows that are entirely -1\n")
  data_matrix_kmeans <- data_matrix_kmeans[!rows_all_neg1, ]
  decoy_ids_kmeans <- decoy_ids_kmeans[!rows_all_neg1]
  rownames(data_matrix_kmeans) <- decoy_ids_kmeans
}

# Transpose matrix: species become rows (y-axis), decoy IDs become columns (x-axis)
data_matrix_kmeans <- t(data_matrix_kmeans)
colnames(data_matrix_kmeans) <- decoy_ids_kmeans

# Map species IDs to common names for row labels
species_id_to_common <- setNames(species_mapping$matched_common_name, species_mapping$species_id)
current_rownames_kmeans <- rownames(data_matrix_kmeans)
common_names_kmeans <- ifelse(current_rownames_kmeans %in% names(species_id_to_common),
  species_id_to_common[current_rownames_kmeans],
  current_rownames_kmeans
)
rownames(data_matrix_kmeans) <- common_names_kmeans

# Final check for kmeans matrix
cat("\nK-means matrix final check:\n")
cat("  Matrix dimensions:", dim(data_matrix_kmeans), "\n")
cat("  -1 values:", sum(data_matrix_kmeans == -1), "\n")
cat("  Value range:", min(data_matrix_kmeans), "to", max(data_matrix_kmeans), "\n")
```

## Create Color Palette and K-means Clustering on Columns (Decoy IDs)

```{r color-paletteandclustering}
# Create color palette: grey for -1, viridis for 0-1
n_colors <- 100

# Use grey for -1, reversed viridis palette for 0-1 range (yellow to purple)
color_gradient <- c("grey70", viridis(n_colors))

# Perform kmeans clustering on columns (decoy IDs) with 6 centers
data_for_kmeans <- t(data_matrix_kmeans)

# Perform kmeans clustering (k=6)
set.seed(42) # For reproducibility
kmeans_result <- kmeans(data_for_kmeans, centers = 6, iter.max = 100, nstart = 10)

# Get cluster assignments
cluster_assignments <- kmeans_result$cluster

# Sort columns by cluster assignment, then by ascending max value within each cluster
col_max_values <- apply(data_matrix_kmeans, 2, function(x) {
  x_clean <- x[x != -1] # Remove -1 values
  if (length(x_clean) == 0) {
    return(-1)
  } # If all -1, return -1
  max(x_clean, na.rm = TRUE)
})

# Create a data frame with column index, cluster assignment, and max value
col_info <- data.frame(
  col_idx = seq_len(ncol(data_matrix_kmeans)),
  cluster = cluster_assignments,
  max_val = col_max_values
)

# Sort: first by cluster, then by ascending max value within cluster
col_info_sorted <- col_info[order(col_info$cluster, col_info$max_val), ]
cluster_order <- col_info_sorted$col_idx

# Reorder columns based on cluster and max value sorting
data_matrix_kmeans_clustered <- data_matrix_kmeans[, cluster_order, drop = FALSE]

# Ensure column names are preserved (decoy IDs)
colnames(data_matrix_kmeans_clustered) <- colnames(data_matrix_kmeans)[cluster_order]

# Create column annotation data frame with cluster numbers (matching the reordered columns)
col_annotation <- data.frame(
  Cluster = factor(col_info_sorted$cluster),
  row.names = colnames(data_matrix_kmeans_clustered)
)

cat("K-means clustering completed:\n")
cat("  Number of clusters: 6\n")
cat("  Cluster sizes:\n")
for (i in 1:6) {
  cat(sprintf("    Cluster %d: %d decoy IDs\n", i, sum(cluster_assignments == i)))
}
cat("  Columns reordered by cluster assignment, then by ascending max value within each cluster\n")
```

## Generate Heatmap with K-means Clustering

{r heatmap-kmeans, fig.width=20, fig.height=16}
# Create heatmap with pre-clustered columns using kmeans (k=6)
png("CDS_Decoy_SpliceAI_MSA_Heatmapgrey70.png", width = 2400, height = 2000, res = 150)
pheatmap(
  data_matrix_kmeans_clustered,
  color = color_gradient,
  na_col = "grey70", # For any remaining NA values
  cluster_rows = FALSE, # Keep species in Newick order
  cluster_cols = FALSE, # Use pre-computed kmeans order
  show_rownames = TRUE,
  show_colnames = FALSE,
  fontsize_row = 8, # Font size for species names
  fontsize_col = 6,
  main = "Coding Sequence Decoy Site SpliceAI Scores Across 100 Species (K-means k=6, -1 = Missing)",
  treeheight_row = 0, # No row dendrogram (using fixed order)
  treeheight_col = 0, # No column dendrogram (using kmeans order)
  breaks = c(-1.1, -0.5, seq(0, 1, length.out = n_colors)), # Grey for -1, viridis for 0-1
  annotation_col = col_annotation # Add cluster number annotations to columns
)
dev.off()
beep()

# Control Dataset Analysis

## Load Control Data

```{r load-control-data}
# Read the filtered control TSV file with suffixes
control_data <- fread("control_spliceai_filtered_protein_coding.tsv", sep = "\t", header = TRUE)
cat("Control data loaded:\n")
cat("  Number of decoy IDs:", nrow(control_data), "\n")
cat("  Number of species:", ncol(control_data) - 1, "\n")

# Separate C (canonical) and P (non-PRPF8) data
control_data_C <- control_data[grepl("_C$", control_data$decoyID), ]
control_data_P <- control_data[grepl("_P$", control_data$decoyID), ]

cat("\nData split:\n")
cat("  Canonical (_C):", nrow(control_data_C), "decoy IDs\n")
cat("  Non-PRPF8 (_P):", nrow(control_data_P), "decoy IDs\n")
```

## Prepare Control Data Matrices for K-means Clustering (C and P separately)

```{r prepare-control-matrix-kmeans}
# Function to prepare matrix for a subset
prepare_control_subset <- function(subset_data, subset_name) {
  # Extract decoy IDs (first column)
  decoy_ids <- subset_data$decoyID
  
  # Extract numeric data (all columns except first)
  data_matrix <- as.matrix(subset_data[, -1, with = FALSE])
  
  # Set row names
  rownames(data_matrix) <- decoy_ids
  
  # Reorder columns to match Newick order (reuse from main section)
  data_matrix <- data_matrix[, ordered_species_ids, drop = FALSE]
  
  # Remove rows that are entirely -1 (can't cluster these)
  rows_all_neg1 <- rowSums(data_matrix != -1) == 0
  if (sum(rows_all_neg1) > 0) {
    cat(paste0(subset_name, " K-means matrix: Removing ", sum(rows_all_neg1), " rows that are entirely -1\n"))
    data_matrix <- data_matrix[!rows_all_neg1, ]
    decoy_ids <- decoy_ids[!rows_all_neg1]
    rownames(data_matrix) <- decoy_ids
  }
  
  # Transpose matrix: species become rows (y-axis), decoy IDs become columns (x-axis)
  data_matrix <- t(data_matrix)
  colnames(data_matrix) <- decoy_ids
  
  # Map species IDs to common names for row labels (reuse mapping from main section)
  current_rownames <- rownames(data_matrix)
  common_names <- ifelse(current_rownames %in% names(species_id_to_common),
    species_id_to_common[current_rownames],
    current_rownames
  )
  rownames(data_matrix) <- common_names
  
  cat(paste0("\n", subset_name, " K-means matrix final check:\n"))
  cat("  Matrix dimensions:", dim(data_matrix), "\n")
  cat("  -1 values:", sum(data_matrix == -1), "\n")
  cat("  Value range:", min(data_matrix), "to", max(data_matrix), "\n")
  
  return(data_matrix)
}

# Prepare C and P matrices separately
control_data_matrix_C <- prepare_control_subset(control_data_C, "Control C")
control_data_matrix_P <- prepare_control_subset(control_data_P, "Control P")
```

## Create Color Palette and K-means Clustering on Control Columns (C and P separately)

```{r color-paletteandclustering-control}
# Function to perform k-means clustering and ordering for a subset
cluster_and_order_control <- function(data_matrix, subset_name) {
  # Perform kmeans clustering on columns (decoy IDs) with 6 centers
  data_for_kmeans <- t(data_matrix)
  
  # Perform kmeans clustering (k=6)
  set.seed(42) # For reproducibility
  kmeans_result <- kmeans(data_for_kmeans, centers = 6, iter.max = 100, nstart = 10)
  
  # Get cluster assignments
  cluster_assignments <- kmeans_result$cluster
  
  # Sort columns by cluster assignment, then by ascending max value within each cluster
  # First, calculate max value for each column (ignoring -1 values)
  col_max_values <- apply(data_matrix, 2, function(x) {
    x_clean <- x[x != -1] # Remove -1 values
    if (length(x_clean) == 0) {
      return(-1)
    } # If all -1, return -1
    max(x_clean, na.rm = TRUE)
  })
  
  # Create a data frame with column index, cluster assignment, and max value
  col_info <- data.frame(
    col_idx = seq_len(ncol(data_matrix)),
    cluster = cluster_assignments,
    max_val = col_max_values
  )
  
  # Sort: first by cluster, then by ascending max value within cluster
  col_info_sorted <- col_info[order(col_info$cluster, col_info$max_val), ]
  cluster_order <- col_info_sorted$col_idx
  
  # Reorder columns based on cluster and max value sorting
  data_matrix_clustered <- data_matrix[, cluster_order, drop = FALSE]
  
  # Ensure column names are preserved (decoy IDs)
  colnames(data_matrix_clustered) <- colnames(data_matrix)[cluster_order]
  
  # Create column annotation data frame with cluster numbers (matching the reordered columns)
  col_annotation <- data.frame(
    Cluster = factor(col_info_sorted$cluster),
    row.names = colnames(data_matrix_clustered)
  )
  
  cat(paste0(subset_name, " K-means clustering completed:\n"))
  cat("  Number of clusters: 6\n")
  cat("  Cluster sizes:\n")
  for (i in 1:6) {
    cat(sprintf("    Cluster %d: %d decoy IDs\n", i, sum(cluster_assignments == i)))
  }
  cat("  Columns reordered by cluster assignment, then by ascending max value within each cluster\n")
  
  return(list(matrix = data_matrix_clustered, annotation = col_annotation))
}

# Cluster and order both C and P subsets
control_C_result <- cluster_and_order_control(control_data_matrix_C, "Control C")
control_P_result <- cluster_and_order_control(control_data_matrix_P, "Control P")

control_data_matrix_C_clustered <- control_C_result$matrix
control_col_annotation_C <- control_C_result$annotation

control_data_matrix_P_clustered <- control_P_result$matrix
control_col_annotation_P <- control_P_result$annotation
```

## Extract Clusters 2 and 4 from Canonical Control and Match with Decoy/NonPRPF8

```{r load-phylop-data}
# Load phyloP data for annotation
# Load filtered phyloP summary data (decoy) - protein coding filtered
phylop_data <- read.table("phyloP/decoy/phyloP_decoy_protein_coding.tsv", 
                          sep = "\t", 
                          header = TRUE,
                          stringsAsFactors = FALSE)

cat("Loaded phyloP decoy data (protein coding filtered):\n")
cat("  Total rows:", nrow(phylop_data), "\n")
cat("  Columns:", paste(colnames(phylop_data), collapse = ", "), "\n")

# Load filtered phyloP control data
phylop_control_data <- fread("phyloP/ctrl/phyloP_ctrl_summary_filtered_protein_coding.tsv", 
                             sep = "\t", 
                             header = TRUE)

cat("\nLoaded phyloP control data:\n")
cat("  Total rows:", nrow(phylop_control_data), "\n")
cat("  Columns:", paste(colnames(phylop_control_data), collapse = ", "), "\n")

# Prepare phyloP data for merging
# Add _D suffix to decoy phyloP data
phylop_decoy_merged <- phylop_data %>%
  mutate(decoyID = paste0(MAF_file, "_D")) %>%
  select(decoyID, mean_phylop, max_phylop, fraction_constrained, 
         fraction_constrained_p01, sum_positive_phylop)

# Control phyloP data already has decoyIDs with suffixes
phylop_control_merged <- phylop_control_data %>%
  select(decoyID, mean_phylop, max_phylop, fraction_constrained, 
         fraction_constrained_p01, sum_positive_phylop)

# Combine phyloP data
phylop_all <- bind_rows(phylop_decoy_merged, phylop_control_merged) %>%
  mutate(
    mean_phylop = as.numeric(mean_phylop),
    max_phylop = as.numeric(max_phylop),
    fraction_constrained = as.numeric(fraction_constrained),
    fraction_constrained_p01 = as.numeric(fraction_constrained_p01),
    sum_positive_phylop = as.numeric(sum_positive_phylop)
  )

cat("\nCombined phyloP data:\n")
cat("  Total rows:", nrow(phylop_all), "\n")
cat("  Rows with phyloP data:", sum(!is.na(phylop_all$mean_phylop)), "\n")
```

```{r extract-clusters-2-4-combined}
# Function to extract gene name from decoyID
# Format: GENE_COORDINATE_SUFFIX (e.g., "GENE_chr1_12345_67890_C")
extract_gene_name <- function(decoy_id) {
  # Remove suffix if present (_C, _P, _D)
  base_id <- gsub("_(C|P|D)$", "", decoy_id)
  # Extract gene name (everything before the last underscore, which is the coordinate)
  parts <- strsplit(base_id, "_")[[1]]
  if (length(parts) > 1) {
    # Gene name is everything except the last part (coordinate)
    paste(parts[-length(parts)], collapse = "_")
  } else {
    base_id
  }
}

# Step 1: Extract clusters 2 and 4 from Canonical control (_C) only
# These clusters will define our gene list
control_C_cluster2 <- rownames(control_col_annotation_C)[control_col_annotation_C$Cluster == 2]
control_C_cluster4 <- rownames(control_col_annotation_C)[control_col_annotation_C$Cluster == 4]

# Combine all decoyIDs from Canonical clusters 2 and 4
canonical_cluster_2_4_decoyIDs <- c(control_C_cluster2, control_C_cluster4)

cat("Step 1: Extracted clusters 2 and 4 from Canonical control (_C):\n")
cat("  Canonical cluster 2:", length(control_C_cluster2), "decoyIDs\n")
cat("  Canonical cluster 4:", length(control_C_cluster4), "decoyIDs\n")
cat("  Total Canonical decoyIDs:", length(canonical_cluster_2_4_decoyIDs), "\n")

# Step 2: Extract unique gene names from these Canonical decoyIDs
# This gene list will be used to match in Decoy and NonPRPF8 datasets
canonical_cluster_2_4_genes <- unique(sapply(canonical_cluster_2_4_decoyIDs, extract_gene_name))
cat("\nStep 2: Extracted unique gene names from Canonical clusters 2 & 4:\n")
cat("  Unique genes:", length(canonical_cluster_2_4_genes), "\n")

# Step 3: Match these genes in Decoy dataset (_D)
# Extract all decoyIDs from Decoy dataset that belong to these genes
decoy_decoyIDs <- spliceai_data$decoyID
decoy_genes <- sapply(decoy_decoyIDs, extract_gene_name)
decoy_matched <- decoy_decoyIDs[decoy_genes %in% canonical_cluster_2_4_genes]

# Add _D suffix to matched Decoy decoyIDs
decoy_matched_labeled <- paste0(decoy_matched, "_D")

cat("\nStep 3: Matched genes in Decoy dataset (_D):\n")
cat("  Matched decoyIDs:", length(decoy_matched_labeled), "\n")

# Step 4: Match these genes in NonPRPF8 control dataset (_P)
# Extract ALL instances of these genes from NonPRPF8 control (not just specific clusters)
all_control_P_decoyIDs <- colnames(control_data_matrix_P_clustered)
all_control_P_genes <- sapply(all_control_P_decoyIDs, extract_gene_name)

# Find all NonPRPF8 decoyIDs with matching genes
control_P_matched <- all_control_P_decoyIDs[all_control_P_genes %in% canonical_cluster_2_4_genes]

cat("\nStep 4: Matched genes in NonPRPF8 control dataset (_P):\n")
cat("  Matched decoyIDs:", length(control_P_matched), "\n")

# Step 5: For Canonical (_C), use the original cluster 2 and 4 decoyIDs
# (These already have the _C suffix from the control data)
control_C_matched <- canonical_cluster_2_4_decoyIDs

cat("\nStep 5: Canonical control (_C) decoyIDs (from clusters 2 & 4):\n")
cat("  Canonical decoyIDs:", length(control_C_matched), "\n")

# Step 6: Combine all matched decoyIDs from all three datasets
all_matched_decoyIDs <- c(decoy_matched_labeled, control_C_matched, control_P_matched)

cat("\nStep 6: Combined all matched decoyIDs:\n")
cat("  Total decoyIDs:", length(all_matched_decoyIDs), "\n")
cat("    Decoy (_D):", length(decoy_matched_labeled), "\n")
cat("    Canonical (_C):", length(control_C_matched), "\n")
cat("    NonPRPF8 (_P):", length(control_P_matched), "\n")
```

```{r prepare-combined-heatmap-clusters-2-4}
# Prepare matrices for combined heatmap
# This will combine Decoy (_D), Canonical (_C), and NonPRPF8 (_P) data

# Extract Decoy data (_D)
decoy_matched_data <- spliceai_data[decoy_decoyIDs %in% decoy_matched, ]
decoy_matched_data$decoyID <- decoy_matched_labeled

# Extract species columns
species_cols <- setdiff(colnames(decoy_matched_data), "decoyID")

# Reorder columns to match Newick order (reuse from main section)
decoy_matrix <- as.matrix(decoy_matched_data[, ..species_cols])
rownames(decoy_matrix) <- decoy_matched_labeled
decoy_matrix <- decoy_matrix[, ordered_species_ids, drop = FALSE]

# Transpose for heatmap (species as rows, decoyIDs as columns)
decoy_matrix_t <- t(decoy_matrix)

# Extract Canonical control data (_C) - from clusters 2 and 4
control_C_matrix <- control_data_matrix_C_clustered[, control_C_matched, drop = FALSE]

# Extract NonPRPF8 control data (_P) - all instances of matched genes
control_P_matrix <- control_data_matrix_P_clustered[, control_P_matched, drop = FALSE]

# Combine all three matrices (columns are decoyIDs with suffixes _D, _C, _P)
combined_matrix <- cbind(decoy_matrix_t, control_C_matrix, control_P_matrix)

# Map species IDs to common names for row labels
current_rownames_combined <- rownames(combined_matrix)
common_names_combined <- ifelse(current_rownames_combined %in% names(species_id_to_common),
  species_id_to_common[current_rownames_combined],
  current_rownames_combined
)
rownames(combined_matrix) <- common_names_combined

cat("\nCombined matrix prepared:\n")
cat("  Dimensions:", dim(combined_matrix), "\n")
cat("  Species (rows):", nrow(combined_matrix), "\n")
cat("  DecoyIDs (columns):", ncol(combined_matrix), "\n")

# Calculate max scores for sorting
col_max_values_combined <- apply(combined_matrix, 2, function(x) {
  x_clean <- x[x != -1]
  if (length(x_clean) == 0) {
    return(-1)
  }
  max(x_clean, na.rm = TRUE)
})

# Create origin annotation based on suffix (_D = Decoy, _C = Canonical, _P = NonPRPF8)
origin_annotation <- data.frame(
  Origin = factor(ifelse(grepl("_D$", colnames(combined_matrix)), "Decoy",
                  ifelse(grepl("_C$", colnames(combined_matrix)), "Canonical",
                  ifelse(grepl("_P$", colnames(combined_matrix)), "NonPRPF8", "Unknown")))),
  row.names = colnames(combined_matrix)
)

# Sort by origin, then by max score within each origin
combined_df_sort <- data.frame(
  decoyID = colnames(combined_matrix),
  origin = origin_annotation$Origin,
  max_score = col_max_values_combined
)
combined_df_sort <- combined_df_sort[order(combined_df_sort$origin, combined_df_sort$max_score), ]

# Reorder matrix columns
combined_matrix_sorted <- combined_matrix[, combined_df_sort$decoyID, drop = FALSE]

# Update annotation to match sorted order
origin_annotation_sorted <- data.frame(
  Origin = combined_df_sort$origin,
  row.names = combined_df_sort$decoyID
)

cat("\nSorted by origin and max score:\n")
print(table(origin_annotation_sorted$Origin))

# Add phyloP annotations
# Merge phyloP data with the sorted decoyIDs
phylop_annotation_sorted <- combined_df_sort %>%
  left_join(phylop_all, by = "decoyID") %>%
  select(decoyID, mean_phylop, max_phylop, fraction_constrained, 
         fraction_constrained_p01, sum_positive_phylop)

# Set row names (using base R approach)
rownames(phylop_annotation_sorted) <- phylop_annotation_sorted$decoyID
phylop_annotation_sorted$decoyID <- NULL

# Replace NA with 0 or create categorical annotations
# For heatmap annotations, we'll use sum_positive_phylop
phylop_annotation_sorted <- phylop_annotation_sorted %>%
  mutate(
    sum_positive_phylop_clean = ifelse(is.na(sum_positive_phylop) | sum_positive_phylop <= 0, 0, sum_positive_phylop)
  )

cat("\nphyloP annotation added:\n")
cat("  DecoyIDs with phyloP data:", sum(!is.na(phylop_annotation_sorted$sum_positive_phylop)), "\n")
cat("  Sum positive phyloP range:", min(phylop_annotation_sorted$sum_positive_phylop_clean, na.rm = TRUE), 
    "to", max(phylop_annotation_sorted$sum_positive_phylop_clean, na.rm = TRUE), "\n")

# Apply log transformation for color scaling (using log1p to handle zeros: log(1+x))
# This will use natural log (ln) scaling for the color annotation
phylop_annotation_sorted <- phylop_annotation_sorted %>%
  mutate(
    sum_positive_phylop_log = log1p(sum_positive_phylop_clean)
  )

cat("  Log-transformed phyloP range:", min(phylop_annotation_sorted$sum_positive_phylop_log, na.rm = TRUE), 
    "to", max(phylop_annotation_sorted$sum_positive_phylop_log, na.rm = TRUE), "\n")

# Combine origin and phyloP annotations (using log-transformed values for color scaling)
combined_annotation <- data.frame(
  Origin = origin_annotation_sorted$Origin,
  Sum_positive_phyloP = phylop_annotation_sorted$sum_positive_phylop_log,
  row.names = rownames(origin_annotation_sorted)
)
```

{r heatmap-combined-clusters-2-4, fig.width=20, fig.height=16}
# Create combined heatmap showing:
# - Canonical (_C): decoyIDs from clusters 2 & 4 only
# - Decoy (_D): all decoyIDs from genes found in Canonical clusters 2 & 4
# - NonPRPF8 (_P): all decoyIDs from genes found in Canonical clusters 2 & 4
# Sorted by origin (D, C, P) then by max SpliceAI score within each origin segment
# Includes annotations: Origin (Decoy/Canonical/NonPRPF8) and phyloP metrics

png("Combined_Clusters2_4_Heatmap_Origin_phyloP_lnAnnotation0126.png", width = 2400, height = 2000, res = 150)
pheatmap(
  combined_matrix_sorted,
  color = color_gradient,
  na_col = "grey90",
  cluster_rows = FALSE, # Keep species in Newick order
  cluster_cols = FALSE, # Use pre-sorted order (by origin and max score)
  show_rownames = TRUE,
  show_colnames = FALSE,
  fontsize_row = 8,
  fontsize_col = 6,
  main = "Combined Heatmap: Genes from Canonical Clusters 2 & 4",
  subtitle = "Canonical (_C): clusters 2 & 4 only | Decoy (_D) & NonPRPF8 (_P): all matching genes | Sorted by origin (D, C, P) and max score | Annotations: Origin + Sum positive phyloP (ln-scaled)",
  treeheight_row = 0,
  treeheight_col = 0,
  breaks = c(-1.1, -0.5, seq(0, 1, length.out = n_colors)),
  annotation_col = combined_annotation,
  annotation_colors = list(
    Origin = c("Decoy" = "salmon", "Canonical" = "darkgreen", "NonPRPF8" = "steelblue"),
    Sum_positive_phyloP = viridis(100)
  )
)
dev.off()
beep()

cat("\nHeatmap saved: Combined_Clusters2_4_Heatmap_Origin_phyloP_Annotation.png\n")
cat("  Shows genes from Canonical clusters 2 & 4 across all three datasets (D, C, P)\n")
cat("  Includes annotations: Origin and Sum positive phyloP conservation metric (ln-scaled for color)\n")


## Extract Top 50% of Decoys by Max SpliceAI Score

```{r save-top50-decoys}
# Extract Decoy data points from the combined dataset
# These are the decoyIDs with _D suffix that matched genes from Canonical clusters 2 & 4
decoy_data_for_export <- decoy_matched_data

# Calculate max SpliceAI score for each Decoy (ignoring -1 values)
decoy_max_scores <- apply(decoy_data_for_export[, ..species_cols], 1, function(x) {
  x_clean <- x[x != -1 & !is.na(x)]
  if (length(x_clean) == 0) {
    return(-1)
  }
  max(x_clean, na.rm = TRUE)
})

# Add max score column to the data
decoy_data_for_export$max_spliceai_score <- decoy_max_scores

# Remove rows with max score of -1 (all missing data)
decoy_data_for_export <- decoy_data_for_export[decoy_data_for_export$max_spliceai_score != -1, ]

# Order by max SpliceAI score (descending)
decoy_data_for_export <- decoy_data_for_export[order(-decoy_data_for_export$max_spliceai_score), ]

# Calculate top 50% threshold
n_total <- nrow(decoy_data_for_export)
n_top50 <- ceiling(n_total * 0.5)

cat("\nDecoy data summary:\n")
cat("  Total Decoy decoyIDs:", n_total, "\n")
cat("  Top 50% count:", n_top50, "\n")
cat("  Max score range:", min(decoy_data_for_export$max_spliceai_score), 
    "to", max(decoy_data_for_export$max_spliceai_score), "\n")
cat("  Top 50% max score threshold:", decoy_data_for_export$max_spliceai_score[n_top50], "\n")

# Extract top 50%
top50_decoys <- decoy_data_for_export[1:n_top50, ]

# Save to TSV file
output_file <- "Top50_Decoys_Clusters2_4_MaxSpliceAI.tsv"
fwrite(top50_decoys, file = output_file, sep = "\t")

cat("\nTop 50% of Decoys saved to:", output_file, "\n")
cat("  Contains", nrow(top50_decoys), "decoyIDs ordered by max SpliceAI score\n")
cat("  Max score range in top 50%:", min(top50_decoys$max_spliceai_score), 
    "to", max(top50_decoys$max_spliceai_score), "\n")
```

## Generate Faceted Control Heatmap with K-means Clustering (P left, C right)

{r heatmap-control-kmeans-faceted, fig.width=32, fig.height=16}
# Create faceted heatmap: P (non-PRPF8) on left, C (canonical) on right
png("CDS_Faceted_Control_SpliceAI_MSA_heatmap.png", width = 4000, height = 2000, res = 150)

# Create left panel: P (non-PRPF8)
p_P <- pheatmap(
  control_data_matrix_P_clustered,
  color = color_gradient,
  na_col = "grey70", # For any remaining NA values
  cluster_rows = FALSE, # Keep species in Newick order
  cluster_cols = FALSE, # Use pre-computed kmeans order
  show_rownames = TRUE,
  show_colnames = FALSE,
  fontsize_row = 8, # Font size for species names
  fontsize_col = 6,
  main = "Non-PRPF8 Sites (K-means k=6, -1 = Missing)",
  treeheight_row = 0, # No row dendrogram (using fixed order)
  treeheight_col = 0, # No column dendrogram (using kmeans order)
  breaks = c(-1.1, -0.5, seq(0, 1, length.out = n_colors)), # Grey for -1, viridis for 0-1
  annotation_col = control_col_annotation_P, # Add cluster number annotations to columns
  silent = TRUE
)

# Create right panel: C (canonical)
p_C <- pheatmap(
  control_data_matrix_C_clustered,
  color = color_gradient,
  na_col = "grey70", # For any remaining NA values
  cluster_rows = FALSE, # Keep species in Newick order
  cluster_cols = FALSE, # Use pre-computed kmeans order
  show_rownames = TRUE,
  show_colnames = FALSE,
  fontsize_row = 8, # Font size for species names
  fontsize_col = 6,
  main = "Canonical Sites (K-means k=6, -1 = Missing)",
  treeheight_row = 0, # No row dendrogram (using fixed order)
  treeheight_col = 0, # No column dendrogram (using kmeans order)
  breaks = c(-1.1, -0.5, seq(0, 1, length.out = n_colors)), # Grey for -1, viridis for 0-1
  annotation_col = control_col_annotation_C, # Add cluster number annotations to columns
  silent = TRUE
)

# Arrange side by side: P on left, C on right
grid.arrange(
  p_P$gtable,
  p_C$gtable,
  ncol = 2,
  widths = c(ncol(control_data_matrix_P_clustered), ncol(control_data_matrix_C_clustered))
)

dev.off()
beep()

# Unstranded Overlaps Dataset Analysis

## Load Unstranded Overlaps Data

```{r load-unstranded-overlaps-data}
# Read the unstranded overlaps TSV file
spliceai_unstranded_data <- fread("SpliceAI_MAF_non_exonic_unstranded_overlaps.tsv", sep = "\t", header = TRUE)
cat("Unstranded overlaps dataset loaded:\n")
cat("  Number of decoy IDs:", nrow(spliceai_unstranded_data), "\n")
cat("  Number of species:", ncol(spliceai_unstranded_data) - 1, "\n")
```

## Prepare Unstranded Overlaps Data Matrix for K-means Clustering

```{r prepare-unstranded-matrix-kmeans}
# Extract decoy IDs (first column)
decoy_ids_unstranded_kmeans <- spliceai_unstranded_data$decoyID

# Extract numeric data (all columns except first)
data_matrix_unstranded_kmeans <- as.matrix(spliceai_unstranded_data[, -1, with = FALSE])

# Set row names
rownames(data_matrix_unstranded_kmeans) <- decoy_ids_unstranded_kmeans

# Reorder columns to match Newick order (reuse from main section)
data_matrix_unstranded_kmeans <- data_matrix_unstranded_kmeans[, ordered_species_ids, drop = FALSE]

# Remove rows that are entirely -1 (can't cluster these)
rows_all_neg1_unstranded <- rowSums(data_matrix_unstranded_kmeans != -1) == 0
if (sum(rows_all_neg1_unstranded) > 0) {
  cat("Unstranded overlaps K-means matrix: Removing", sum(rows_all_neg1_unstranded), "rows that are entirely -1\n")
  data_matrix_unstranded_kmeans <- data_matrix_unstranded_kmeans[!rows_all_neg1_unstranded, ]
  decoy_ids_unstranded_kmeans <- decoy_ids_unstranded_kmeans[!rows_all_neg1_unstranded]
  rownames(data_matrix_unstranded_kmeans) <- decoy_ids_unstranded_kmeans
}

# Transpose matrix: species become rows (y-axis), decoy IDs become columns (x-axis)
data_matrix_unstranded_kmeans <- t(data_matrix_unstranded_kmeans)
colnames(data_matrix_unstranded_kmeans) <- decoy_ids_unstranded_kmeans

# Map species IDs to common names for row labels (reuse mapping from main section)
current_rownames_unstranded_kmeans <- rownames(data_matrix_unstranded_kmeans)
common_names_unstranded_kmeans <- ifelse(current_rownames_unstranded_kmeans %in% names(species_id_to_common),
  species_id_to_common[current_rownames_unstranded_kmeans],
  current_rownames_unstranded_kmeans
)
rownames(data_matrix_unstranded_kmeans) <- common_names_unstranded_kmeans

# Final check for unstranded kmeans matrix
cat("\nUnstranded overlaps K-means matrix final check:\n")
cat("  Matrix dimensions:", dim(data_matrix_unstranded_kmeans), "\n")
cat("  -1 values:", sum(data_matrix_unstranded_kmeans == -1), "\n")
cat("  Value range:", min(data_matrix_unstranded_kmeans), "to", max(data_matrix_unstranded_kmeans), "\n")
```

## Create Color Palette and K-means Clustering on Unstranded Overlaps Columns

```{r color-paletteandclustering-unstranded}
# Perform kmeans clustering on columns (decoy IDs) with 6 centers
data_for_kmeans_unstranded <- t(data_matrix_unstranded_kmeans)

# Perform kmeans clustering (k=6)
set.seed(42) # For reproducibility
kmeans_result_unstranded <- kmeans(data_for_kmeans_unstranded, centers = 6, iter.max = 100, nstart = 10)

# Get cluster assignments
cluster_assignments_unstranded <- kmeans_result_unstranded$cluster

# Sort columns by cluster assignment, then by ascending max value within each cluster
col_max_values_unstranded <- apply(data_matrix_unstranded_kmeans, 2, function(x) {
  x_clean <- x[x != -1] # Remove -1 values
  if (length(x_clean) == 0) {
    return(-1)
  } # If all -1, return -1
  max(x_clean, na.rm = TRUE)
})

# Create a data frame with column index, cluster assignment, and max value
col_info_unstranded <- data.frame(
  col_idx = seq_len(ncol(data_matrix_unstranded_kmeans)),
  cluster = cluster_assignments_unstranded,
  max_val = col_max_values_unstranded
)

# Sort: first by cluster, then by ascending max value within cluster
col_info_sorted_unstranded <- col_info_unstranded[order(col_info_unstranded$cluster, col_info_unstranded$max_val), ]
cluster_order_unstranded <- col_info_sorted_unstranded$col_idx

# Reorder columns based on cluster and max value sorting
data_matrix_unstranded_kmeans_clustered <- data_matrix_unstranded_kmeans[, cluster_order_unstranded, drop = FALSE]

# Ensure column names are preserved (decoy IDs)
colnames(data_matrix_unstranded_kmeans_clustered) <- colnames(data_matrix_unstranded_kmeans)[cluster_order_unstranded]

# Create column annotation data frame with cluster numbers
col_annotation_unstranded <- data.frame(
  Cluster = factor(col_info_sorted_unstranded$cluster),
  row.names = colnames(data_matrix_unstranded_kmeans_clustered)
)

cat("Unstranded overlaps K-means clustering completed:\n")
cat("  Number of clusters: 6\n")
cat("  Cluster sizes:\n")
for (i in 1:6) {
  cat(sprintf("    Cluster %d: %d decoy IDs\n", i, sum(cluster_assignments_unstranded == i)))
}
cat("  Columns reordered by cluster assignment, then by ascending max value within each cluster\n")
```

## Generate Heatmap for Unstranded Overlaps with K-means Clustering

{r heatmap-unstranded-kmeans, fig.width=20, fig.height=16}
# Create heatmap with pre-clustered columns using kmeans (k=6)
# Matrix is transposed: species are rows (y-axis), decoy IDs are columns (x-axis)
png("UnstrandedOverlaps_SpliceAI_MSA_Heatmap.png", width = 2400, height = 2000, res = 150)
pheatmap(
  data_matrix_unstranded_kmeans_clustered,
  color = color_gradient,
  na_col = "grey70", # For any remaining NA values
  cluster_rows = FALSE, # Keep species in Newick order
  cluster_cols = FALSE, # Use pre-computed kmeans order
  show_rownames = TRUE,
  show_colnames = FALSE,
  fontsize_row = 8, # Font size for species names
  fontsize_col = 6,
  main = "Unstranded Overlaps Decoy Site SpliceAI Scores Across 100 Species (K-means k=6, -1 = Missing)",
  treeheight_row = 0, # No row dendrogram (using fixed order)
  treeheight_col = 0, # No column dendrogram (using kmeans order)
  breaks = c(-1.1, -0.5, seq(0, 1, length.out = n_colors)), # Grey for -1, viridis for 0-1
  annotation_col = col_annotation_unstranded # Add cluster number annotations to columns
)
dev.off()
beep()

~~~~~~~~~OLD~~~~~~~~~~~~~~OLD~~~~~~~~~~~~~~~~~~~OLD~~~~~~~~~~~~~~OLD~~~~~~~~~~~~~~~~~~OLD~~~~~~~~~~~~~~OLD~~~~~~~~~~~~~~~~~~OLD~~~~~~~~~~~~~~OLD~~~~~

## Prepare Old version of Data Matrix

{r prepare-matrix}
# Extract decoy IDs (first column)
decoy_ids <- spliceai_data$decoyID

# Extract numeric data (all columns except first)
data_matrix <- as.matrix(spliceai_data[, -1, with = FALSE])

# Set row names
rownames(data_matrix) <- decoy_ids

# Check for -1 values (missing species)
cat("Number of -1 values:", sum(data_matrix == -1), "\n")
cat("Percentage of -1 values:", round(100 * sum(data_matrix == -1) / length(data_matrix), 2), "%\n")

# Convert -1 to NA for masking in heatmap
data_matrix[data_matrix == -1] <- NA

# Reorder columns to match Newick order (simplified - we always have 100 species)
data_matrix <- data_matrix[, ordered_species_ids, drop = FALSE]

# Remove rows that are entirely NA (can't cluster these)
rows_all_na <- rowSums(!is.na(data_matrix)) == 0
if (sum(rows_all_na) > 0) {
  cat("Removing", sum(rows_all_na), "rows that are entirely NA\n")
  data_matrix <- data_matrix[!rows_all_na, ]
  decoy_ids <- decoy_ids[!rows_all_na]
  rownames(data_matrix) <- decoy_ids[!rows_all_na]
}

# Transpose matrix: species become rows (y-axis), decoy IDs become columns (x-axis)
data_matrix <- t(data_matrix)

# After transposition, remove rows (species) that are entirely NA
rows_all_na <- rowSums(!is.na(data_matrix)) == 0
if (sum(rows_all_na) > 0) {
  cat("After transposition: Removing", sum(rows_all_na), "rows (species) that are entirely NA\n")
  data_matrix <- data_matrix[!rows_all_na, ]
}

# Remove columns (decoy IDs) that are entirely NA
cols_all_na <- colSums(!is.na(data_matrix)) == 0
if (sum(cols_all_na) > 0) {
  cat("After transposition: Removing", sum(cols_all_na), "columns (decoy IDs) that are entirely NA\n")
  data_matrix <- data_matrix[, !cols_all_na]
}

# Map species IDs to common names for row labels
species_id_to_common <- setNames(species_mapping$matched_common_name, species_mapping$species_id)

# Replace row names (species IDs) with common names
current_rownames <- rownames(data_matrix)
common_names <- ifelse(current_rownames %in% names(species_id_to_common),
  species_id_to_common[current_rownames],
  current_rownames
)
rownames(data_matrix) <- common_names

# Final check: Verify no Inf, NaN, or all-NA rows/columns before kmeans
cat("\nFinal data check before kmeans:\n")
cat("  Matrix dimensions:", dim(data_matrix), "\n")
cat("  Inf values:", sum(is.infinite(data_matrix)), "\n")
cat("  NaN values:", sum(is.nan(data_matrix)), "\n")
cat("  NA values:", sum(is.na(data_matrix)), "\n")
cat("  Rows with all NA:", sum(rowSums(!is.na(data_matrix)) == 0), "\n")
cat("  Columns with all NA:", sum(colSums(!is.na(data_matrix)) == 0), "\n")


## Generate Heatmap with NA Values (Old Version)

{r heatmap, fig.width=20, fig.height=16}
# Create heatmap with clustering (original version with NA values)
# Matrix is transposed: species are rows (y-axis), decoy IDs are columns (x-axis)
png("spliceai_msa_heatmap_orderedspecies.png", width = 2400, height = 2000, res = 150)
pheatmap(
  data_matrix,
  color = color_gradient,
  na_col = "grey40",
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = FALSE,
  fontsize_row = 8, # Font size for species names
  fontsize_col = 6,
  main = "SpliceAI Scores Across 100 Species (Ordered by Newick Tree)",
  treeheight_row = 0, # No row dendrogram (using fixed order)
  treeheight_col = 50 # Height of column dendrogram
)
dev.off()

